### 2024-11-17 paper/paper.qmd
Update paper.qmd

---
title: "My title"
subtitle: "My subtitle if needed"
author: 
  - Manjun Zhu
thanks: "Code and data are available at: [https://github.com/Karrrrmen/Life_Expectancy](https://github.com/Karrrrmen/Life_Expectancy)."
date: today
date-format: long
abstract: "First sentence. Second sentence. Third sentence. Fourth sentence."
format: pdf
number-sections: true
bibliography: references.bib
---

```{r}
#| include: false
#| warning: false
#| message: false

library(dplyr)
library(tidyverse)
library(here)
library(ggplot2)

data <- read.csv(here::here("data/02-analysis_data/analysis_data.csv"))

```

# Introduction

Overview paragraph

Estimand paragraph

Results paragraph

Why it matters paragraph

Telegraphing paragraph: The remainder of this paper is structured as follows. @sec-data....

# Data {#sec-data}

## Overview

We use the statistical programming language R [@citeR].... Our data [@shelter].... Following @tellingstories, we consider...

Overview text

## Measurement

Some paragraphs about how we go from a phenomena in the world to an entry in the dataset.

## Predictor variables

Add graphs, tables and text. Use sub-sub-headings for each outcome variable or update the subheading to be singular.

Some of our data is of WHO (@fig-predictor).

```{r}
#| label: fig-predictor
#| echo: false
#| warning: false

ggplot(data, aes(x = LifeExpectancy)) +
  geom_histogram(binwidth = 1, fill = "purple", color = "brown", alpha = 0.7) +
  facet_wrap(~ Status) +
  labs(
    title = "Histogram of Life Expectancy by Status",
    x = "Life Expectancy",
    y = "Frequency"
  ) +
  theme_minimal()

```

Talk more about it.

And also planes (@fig-planes). (You can change the height and width, but don't worry about doing that until you have finished every other aspect of the paper - Quarto will try to make it look nice and the defaults usually work well once you have enough text.)

```{r}
#| label: fig-planes
#| fig-cap: Relationship between wing length and width
#| echo: false
#| warning: false
#| message: false

```

Talk way more about it.

## Predictor variables

Add graphs, tables and text.

Use sub-sub-headings for each outcome variable and feel free to combine a few into one if they go together naturally.

## Cleaned Data
```{r}
#Overview, delete later
# Randomly select 10 rows from the data
sampled_data <- data |> sample_n(10)

# Display the selected data
sampled_data
```

# Model

The goal of our modelling strategy is twofold. Firstly,...

Here we briefly describe the Bayesian analysis model used to investigate... Background details and diagnostics are included in [Appendix -@sec-model-details].

## Model set-up

Define $y_i$ as the number of seconds that the plane remained aloft. Then $\beta_i$ is the wing width and $\gamma_i$ is the wing length, both measured in millimeters.

\begin{align} 
y_i|\mu_i, \sigma &\sim \mbox{Normal}(\mu_i, \sigma) \\
\mu_i &= \alpha + \beta_i + \gamma_i\\
\alpha &\sim \mbox{Normal}(0, 2.5) \\
\beta &\sim \mbox{Normal}(0, 2.5) \\
\gamma &\sim \mbox{Normal}(0, 2.5) \\
\sigma &\sim \mbox{Exponential}(1)
\end{align}

We run the model in R [@citeR] using the `rstanarm` package of @rstanarm. We use the default priors from `rstanarm`.

### Model justification

We expect a positive relationship between the size of the wings and time spent aloft. In particular...

We can use maths by including latex between dollar signs, for instance $\theta$.

# Results

Our results are summarized in @tbl-modelresults.

```{r}
#| echo: false
#| eval: true
#| warning: false
#| message: false

library(rstanarm)

first_model <-
  readRDS(file = here::here("models/first_model.rds"))
```

```{r}
#| echo: false
#| eval: true
#| label: tbl-modelresults
#| tbl-cap: "Explanatory models of flight time based on wing width and wing length"
#| warning: false

modelsummary::modelsummary(
  list(
    "First model" = first_model
  ),
  statistic = "mad",
  fmt = 2
)
```

# Discussion

## First discussion point {#sec-first-point}

If my paper were 10 pages, then should be be at least 2.5 pages. The discussion is a chance to show off what you know and what you learnt from all this.

## Second discussion point

Please don't use these as sub-heading labels - change them to be what your point actually is.

## Third discussion point

## Weaknesses and next steps

Weaknesses and next steps should also be included.

\newpage

\appendix

# Appendix {-}

# Additional data details

# Model details {#sec-model-details}

## Posterior predictive check

In @fig-ppcheckandposteriorvsprior-1 we implement a posterior predictive check. This shows...

In @fig-ppcheckandposteriorvsprior-2 we compare the posterior with the prior. This shows...

```{r}
#| eval: true
#| echo: false
#| message: false
#| warning: false
#| label: fig-ppcheckandposteriorvsprior
#| layout-ncol: 2
#| fig-cap: "Examining how the model fits, and is affected by, the data"
#| fig-subcap: ["Posterior prediction check", "Comparing the posterior with the prior"]

pp_check(first_model) +
  theme_classic() +
  theme(legend.position = "bottom")

posterior_vs_prior(first_model) +
  theme_minimal() +
  scale_color_brewer(palette = "Set1") +
  theme(legend.position = "bottom") +
  coord_flip()
```

## Diagnostics

@fig-stanareyouokay-1 is a trace plot. It shows... This suggests...

@fig-stanareyouokay-2 is a Rhat plot. It shows... This suggests...

```{r}
#| echo: false
#| eval: true
#| message: false
#| warning: false
#| label: fig-stanareyouokay
#| fig-cap: "Checking the convergence of the MCMC algorithm"
#| fig-subcap: ["Trace plot", "Rhat plot"]
#| layout-ncol: 2

plot(first_model, "trace")

plot(first_model, "rhat")
```

\newpage

# References

### 2024-11-24 scripts/04-test_analysis_data.R
Update 04-test_analysis_data.R

@@ -1,7 +1,7 @@
 #### Preamble ####
 # Purpose: Performs a series of validation tests on cleaned Life Expectancy Data
 # Author: Manjun Zhu
-# Date: 16 November 2024
+# Date: 23 November 2024
 # Contact: karmen.zhu@mail.utoronto.ca
 # License: MIT
 # Pre-requisites: The `tidyverse` , `arrow`, and `testthat` packages must be installed.
@@ -32,7 +32,7 @@ test_results$structure <- test_that("Data structure", {
 # Test if column names match expected names
 test_results$column_names_test <- test_that("Column names", {
   expected_colnames <- c("Country", "Year", "Status", "LifeExpectancy", "GDP", 
-                         "Diphtheria", "IncomeComposition", "TotalExpenditure", "BMI")
+                         "Diphtheria", "Schooling", "TotalExpenditure", "BMI")
   expect_equal(names(analysis_data), expected_colnames, 
                info = "Column names do not match expected names.")
 })
@@ -44,8 +44,8 @@ test_results$na_values_test <- test_that("NA values", {
   expect_false(any(is.na(analysis_data$GDP)), info = "GDP contains NA values.")
   expect_false(any(is.na(analysis_data$Diphtheria)), 
                info = "Diphtheria contains NA values.")
-  expect_false(any(is.na(analysis_data$IncomeComposition)), 
-               info = "IncomeComposition contains NA values.")
+  expect_false(any(is.na(analysis_data$Schooling)), 
+               info = "Schooling contains NA values.")
   expect_false(any(is.na(analysis_data$TotalExpenditure)), 
                info = "TotalExpenditure contains NA values.")
   expect_false(any(is.na(analysis_data$BMI)), info = "BMI contains NA values.")
@@ -95,6 +95,13 @@ test_results$bmi_test <- test_that("BMI range", {
                     analysis_data$BMI <= 50, na.rm = TRUE),
               info = "BMI values are out of expected range (10 to 50).")
 })
+
+# Test if Schooling values are within expected range (3 to 30)
+test_results$schooling_range_test <- test_that("Schooling range", {
+  expect_true(all(analysis_data$Schooling >= 3 & analysis_data$Schooling <= 30), 
+              info = "Schooling values are out of expected range (3 to 30).")
+})
+
 
 # Print results
 print(test_results)

### 2024-11-24 scripts/02-download_data.R
Add supplementary datasets

@@ -1,8 +1,8 @@
 #### Preamble ####
 # Purpose: Downloads and saves the data from 'https://www.kaggle.com/datasets/kumarajarshi/life-expectancy-who'
 # Author: Manjun Zhu
-# Date: 16 November 2024
-# Contact: karmen.zhu@mail.utoronto.ca
+# Date: 24 November 2024
+# Contact: karmen.zhu@utoronto.ca
 # License: University of Toronto
 # Pre-requisites: No
 # Any other information needed? Thanks to Kaggle opendataset
@@ -14,6 +14,28 @@ library(dplyr)
 
 #### Download data ####
 raw_data <- read.csv(file = "data/01-raw_data/LifeExpectancyData.csv")
+
+# Load the supplementary dataset
+doctor <- read.csv("data/01-raw_data/medicalDoctors.csv")
+nursing <- read.csv("data/01-raw_data/nursingAndMidwife.csv")
+
+# Ensure the column names align (modify as needed)
+# Example: rename 'Location' in doctor to 'Country'
+colnames(doctor)[colnames(doctor) == "Location"] <- "Country"
+colnames(doctor)[colnames(doctor) == "Period"] <- "Year"
+colnames(doctor)[colnames(doctor) == "First.Tooltip"] <- "pct_doctor"
+colnames(nursing)[colnames(nursing) == "Location"] <- "Country"
+colnames(nursing)[colnames(nursing) == "Period"] <- "Year"
+colnames(nursing)[colnames(nursing) == "First.Tooltip"] <- "pct_nursing"
+
+# Merge raw_data with doctor dataset
+test_data <- raw_data %>%
+  left_join(doctor %>% select(Country, Year, pct_doctor), by = c("Country", "Year"))
+
+# Merge test_data with nursing dataset
+raw_data <- test_data %>%
+  left_join(nursing %>% select(Country, Year, pct_nursing), by = c("Country", "Year"))
+
 
 #### Save data ####
 # change raw_data to whatever name you assigned when you downloaded it.

### 2024-11-25 scripts/03-clean_data.R
Add additional predictor

@@ -1,7 +1,7 @@
 #### Preamble ####
 # Purpose: Clean the raw data recorded by WHO, removing unnecessary columns, and save it as csv and parquet.
 # Author: Manjun Zhu
-# Date: 16 November 2024
+# Date: 23 November 2024
 # Contact: karmen.zhu@mail.utoronto.ca
 # License: MIT
 # Pre-requisites: finish raw-data
@@ -13,19 +13,20 @@ library(arrow)
 #### Clean data ####
 data <- read_csv("data/01-raw_data/raw_data.csv")
 
-data <- data %>%
+analysis_data <- data %>%
   # Select the desired columns
-  select(Country, Year, Status, Life.expectancy, GDP, Diphtheria, Total.expenditure, BMI, Schooling) %>%
+  select(Country, Year, Status, Life.expectancy, GDP, Diphtheria, 
+         Total.expenditure, BMI, Schooling, pct_doctor, pct_nursing) %>%
   # Filter rows where Year is between 2009 and 2015
-  filter(Year >= 2009 & Year <= 2015) %>%
+  filter(Year >= 2004 & Year <= 2015) %>%
   # Remove rows with any missing values
   drop_na()
 
 #### Rename column
-colnames(data)[colnames(data) == "Life.expectancy"] <- "LifeExpectancy"
-colnames(data)[colnames(data) == 
+colnames(analysis_data)[colnames(analysis_data) == "Life.expectancy"] <- "LifeExpectancy"
+colnames(analysis_data)[colnames(analysis_data) == 
                  "Total.expenditure"] <- "TotalExpenditure"
 
 #### Save data ####
-write_csv(data, "data/02-analysis_data/analysis_data.csv")
-write_parquet(data, "data/02-analysis_data/analysis_data.parquet")
+write_csv(analysis_data, "data/02-analysis_data/analysis_data.csv")
+write_parquet(analysis_data, "data/02-analysis_data/analysis_data.parquet")

### 2024-11-25 scripts/06-model_data.R
change model

@@ -1,11 +1,10 @@
 #### Preamble ####
 # Purpose: Run a linear regression model to predict Life Expectancy between Developed and Developing Countries
 # Author: Manjun Zhu
-# Date: 23 November 2024
+# Date: 24 November 2024
 # Contact: karmen.zhu@mail.utoronto.ca
 # License: MIT
 # Pre-requisites: The `tidyverse` and 'here' packages must be installed
-# Any other information needed? [...UPDATE THIS...]
 
 
 #### Workspace setup ####
@@ -21,24 +20,25 @@ data <- data %>%
     Year = as.numeric(Year),
     LifeExpectancy = as.numeric(LifeExpectancy),
     Status = as.factor(Status),
-    GDP = as.numeric(GDP),
     Diphtheria = as.numeric(Diphtheria),
     BMI = as.numeric(BMI),
     TotalExpenditure = as.numeric(TotalExpenditure),
-    Schooling = as.numeric(Schooling)
+    Schooling = as.numeric(Schooling),
+    pct_doctor = as.numeric(pct_doctor),
+    pct_nursing = as.numeric(pct_nursing)
   )
 
 ### Model data ####
 # linear model for Developed Countries' Life Expectancy based on socioeconomic factors
-lm_developed <- lm(LifeExpectancy ~ GDP + BMI + Diphtheria + Schooling + TotalExpenditure, 
+lm_developed <- lm(LifeExpectancy ~  Diphtheria + Schooling + TotalExpenditure + pct_doctor, 
                    data = data %>% filter(Status == "Developed"))
 
 
 # linear model for Developing Countries' Life Expectancy based on socioeconomic factors
-lm_developing <- lm(LifeExpectancy ~ GDP + BMI + Diphtheria + Schooling + TotalExpenditure,
+lm_developing <- lm(LifeExpectancy ~ BMI + Diphtheria + Schooling + pct_doctor + pct_nursing, 
                     data = data %>% filter(Status == "Developing"))
-
 
 #### Save model ####
 saveRDS(lm_developed, here::here("models/lm_developed.rds"))
 saveRDS(lm_developing, here::here("models/lm_developing.rds"))
+

### 2024-11-27 scripts/03-clean_data.R
Update 03-clean_data.R

@@ -1,7 +1,7 @@
 #### Preamble ####
 # Purpose: Clean the raw data recorded by WHO, removing unnecessary columns, and save it as csv and parquet.
 # Author: Manjun Zhu
-# Date: 23 November 2024
+# Date: 26 November 2024
 # Contact: karmen.zhu@mail.utoronto.ca
 # License: MIT
 # Pre-requisites: finish raw-data
@@ -15,10 +15,12 @@ data <- read_csv("data/01-raw_data/raw_data.csv")
 
 analysis_data <- data %>%
   # Select the desired columns
-  select(Country, Year, Status, Life.expectancy, GDP, Diphtheria, 
+  select(Country, Year, Status, Life.expectancy, Diphtheria, 
          Total.expenditure, BMI, Schooling, pct_doctor, pct_nursing) %>%
-  # Filter rows where Year is between 2009 and 2015
+  # Filter rows where Year is between 2009 and 2015, and drop the unrealistic value
   filter(Year >= 2004 & Year <= 2015) %>%
+  filter(pct_doctor >= 0 & pct_doctor <= 1500) %>%
+  filter(pct_nursing >= 0 & pct_nursing <= 1500) %>%
   # Remove rows with any missing values
   drop_na()
 

### 2024-11-27 scripts/04-test_analysis_data.R
Update 04-test_analysis_data.R

@@ -1,7 +1,7 @@
 #### Preamble ####
 # Purpose: Performs a series of validation tests on cleaned Life Expectancy Data
 # Author: Manjun Zhu
-# Date: 23 November 2024
+# Date: 26 November 2024
 # Contact: karmen.zhu@mail.utoronto.ca
 # License: MIT
 # Pre-requisites: The `tidyverse` , `arrow`, and `testthat` packages must be installed.
@@ -15,6 +15,7 @@ library(arrow)
 
 analysis_data <- read_parquet(here::here("data/02-analysis_data/analysis_data.parquet"))
 
+
 #### Test data ####
 # Create list for results
 test_results <- list()
@@ -23,16 +24,17 @@ test_results <- list()
 test_results$structure <- test_that("Data structure", {
   expect_true(is.data.frame(analysis_data), 
               info = "analysis_data is not a data frame.")
-  expect_equal(ncol(analysis_data), 9, info = "analysis_data does 
+  expect_equal(ncol(analysis_data), 10, info = "analysis_data does 
                not have the correct number of columns.")
-  expect_equal(nrow(analysis_data), 920, info = "analysis_data does 
+  expect_equal(nrow(analysis_data), 1021, info = "analysis_data does 
                not have the correct number of rows.")
 })
 
 # Test if column names match expected names
 test_results$column_names_test <- test_that("Column names", {
   expected_colnames <- c("Country", "Year", "Status", "LifeExpectancy", "GDP", 
-                         "Diphtheria", "Schooling", "TotalExpenditure", "BMI")
+                         "Diphtheria", "TotalExpenditure", "BMI", "Schooling", 
+                         "pct_doctor", "pct_nursing")
   expect_equal(names(analysis_data), expected_colnames, 
                info = "Column names do not match expected names.")
 })
@@ -41,9 +43,12 @@ test_results$column_names_test <- test_that("Column names", {
 test_results$na_values_test <- test_that("NA values", {
   expect_false(any(is.na(analysis_data$LifeExpectancy)), 
                info = "LifeExpectancy contains NA values.")
-  expect_false(any(is.na(analysis_data$GDP)), info = "GDP contains NA values.")
   expect_false(any(is.na(analysis_data$Diphtheria)), 
                info = "Diphtheria contains NA values.")
+  expect_false(any(is.na(analysis_data$pct_doctor)), 
+               info = "pct_doctor contains NA values.")
[…]

### 2024-11-27 paper/paper.qmd
Add files via upload

@@
 ---
-title: "My title"
-subtitle: "My subtitle if needed"
+title: "What to Expect with Life Expectancy?"
+subtitle: "Examining How Status, Healthcare Expenditure, Education, and Medical Infrastructure Influence Life Expectancy Across Developed and Developing Regions"
 author: 
@@
 date-format: long
-abstract: "First sentence. Second sentence. Third sentence. Fourth sentence."
+abstract: "Life expectancy has seen unprecedented growth in recent decades, driven by advancements in healthcare, improved access to medical services, and reduced child mortality.
+However, this growth has not been evenly distributed across the globe, with wealthier nations benefiting disproportionately.
+This research examines the impact of a nation’s economic standing and health conditions on life expectancy, exploring nuanced variables beyond those traditionally studied.
+By developing a linear model, the analysis identifies key relationships between economic and healthcare indicators and their influence on population longevity.
+The findings aim to inform health policy and economic planning, providing actionable insights to address disparities and promote equitable growth in life expectancy worldwide."
 format: pdf
@@
 library(ggplot2)
+library(knitr)
+library(kableExtra)
+library(modelsummary)
+library(ggbeeswarm)
+library(ggridges)
 
@@
 
-Overview paragraph
+Life expectancy has witnessed remarkable growth over the past century, driven by advancements in healthcare, improved access to medical services, and reduced child mortality rates.
+However, this progress has not been distributed equitably across the globe.
+While nations with robust economies and advanced healthcare systems have achieved significant gains, others, particularly in underdeveloped regions, continue to face barriers such as inadequate infrastructure, persistent diseases, and economic instability.
+Despite existing research on life expectancy determinants, gaps remain in understanding the nuanced relationships between economic standing, healthcare access, and other critical variables.
 
-Estimand paragraph
+This study aims to address these gaps by developing a linear model to analyze the impact of a nation’s status, healthcare factors, and economic indicators on life expectancy.
+The response variable is Life Expectancy, while the predictors include Status (Developed vs.
+Developing), Diphtheria Immunization Coverage, Government Total Expenditure, Numbers of Doctors and Nurses per 10000, Schooling and BMI.
+The analysis seeks to quantify the influence of these variables, providing insights into the disparities in life expectancy and identifying factors that contribute to these differences.
 
-Results paragraph
+Previous studies provide valuable context for this research.
+One study [@Garon2015] examined the long-term effects of diphtheria on survivors, revealing that paralyzed individuals experience greater reductions in life expectancy, which underscores the importance of healthcare infrastructure in mitigating preventable diseases in underdeveloped regions.
+Another study extended this understanding by exploring the role of the Human Development Index (HDI), emphasizing the influence of public health initiatives, economic conditions, and education on life expectancy in low and medium-HDI nations [@Girum2018].
+Besides that, advanced econometric techniques, including panel cointegration analysis, have demonstrated a positive relationship between health care expenditure per capita and life expectancy, providing a robust foundation for integrating economic variables into this study [@He2020].
+Additionally, a finding [@Hanushek2016] underscored the importance of combining both educational investments and government support in improving life expectancy, particularly in developing regions.
+The strong correlation between education and health suggests that higher levels of education improve overall health, reducing mortality rates and extending life expectancy.
 
-Why it matters paragraph
+The analysis revealed that life expectancy in developed countries is generally higher and more stable, attributed to stronger healthcare systems, consistent access to medical services, and robust economic stability.
+These factors are typically supported by well-established public health policies.
+In contrast, life expectancy in developing countries demonstrates a broader range of variation, though with a noticeable upward trend.
+This improvement is driven by diverse factors like better healthcare access, economic development, and increasing resource availability, but the inconsistency across regions highlights disparities in infrastructure.
+The faster growth trajectory in these nations indicates progress, yet challenges remain in achieving stable health outcomes.
+These findings suggest that policymakers in developing countries should prioritize healthcare system strengthening, resource access, and reducing regional inequalities, while in developed countries, maintaining consistent healthcare access and supporting sustained economic growth should remain priorities.
 
-Telegraphing paragraph: The remainder of this paper is structured as follows. @sec-data....
+The remainder of this paper is structured as follows: Section @sec-data discusses the data sources and cleaning methods, followed by an analysis of key predictors in Section @sec-model.
+Section @sec-results presents the findings from the linear regression model, and Section @sec-discussion explores the implications of these findings, concluding with recommendations for policy interventions and areas for future research.
 
@@
 
-We use the statistical programming language R [@citeR].... Our data [@shelter].... Following @tellingstories, we consider...
+This study utilizes two primary datasets related to global health and life expectancy.
+The first dataset, World Health Statistics 2020 Complete, sourced from Kaggle [@world_health_statistics], provides comprehensive global health data, including variables such as GDP, life expectancy, mortality rates, and access to healthcare.
+The second dataset, Life Expectancy (WHO), also from Kaggle [@life_expectancy_who], focuses specifically on life expectancy trends and factors, offering a detailed breakdown by country, year, and socioeconomic variables.
 
-Overview text
+Together, these datasets enable an in-depth exploration of the relationships between life expectancy and various predictors, such as economic indicators, healthcare access, and immunization coverage.
+Key variables include Life Expectancy, Status (Developed vs.
+Developing), Diphtheria Immunization Coverage, Government Health Expenditure as percentage of Total Expenditure, Numbers of Doctors and Nurses per 10000, Schooling and BMI.
+
+We performed data cleaning and analysis using R [@citeR], incorporating a range of packages for efficient data manipulation, visualization, and statistical modeling, leveraging packages such as **`tidyverse`** for data manipulation and visualization [@tidyverse], **`dplyr`** for streamlined data wrangling [@dplyr], **`ggplot2`** for creating high-quality plots [@ggplot2], **`ggridges`** for displaying the distribution of variables across different categories with ridge plots [@ggridges], **`here`** for managing file paths [@here], **`lubridate`** for handling date-related variables [@lubridate], **`kableExtra`** for generating polished tables [@kableExtra], **`modelsummary`** for summarizing model results [@modelsummary].
 
@@
 
-Some paragraphs about how we go from a phenomena in the world to an entry in the dataset.
+The measurement process refers to how real-world phenomena—such as healthcare access, economic indicators, or demographic factors—are translated into numerical entries in the dataset.
+Each entry corresponds to a specific country and year, capturing variables related to life expectancy and associated determinants.
+Below is a description of key variables in the dataset:
 
-## Predictor variables
+**Life Expectancy (`LifeExpectancy`)**: This is the dependent variable, representing the average number of years a person is expected to live from birth, assuming current mortality rates.
+The data is reported by international organizations like the World Health Organization (WHO) and is recorded in numeric format as age in years.
 
-Add graphs, tables and text. Use sub-sub-headings for each outcome variable or update the subheading to be singular.
+**Status (`Status`)**: This categorical variable identifies whether a country is classified as "Developed" or "Developing." The classification is based on economic and social indicators defined by the World Health Organization (WHO) or related agencies.
 
-Some of our data is of WHO (@fig-predictor).
+**Body Mass Index (`BMI`)**: This variable captures the average body mass index of the population, calculated as weight in kilograms divided by the square of height in meters.
+It is an aggregate measure derived from population health surveys and recorded as a numeric value.
+
+**Health Expenditure (`TotalExpenditure`)**: This represents the general government expenditure on health as a percentage of total government expenditure.
+It is reported by national governments or international organizations and recorded as a percentage in the dataset.
+
+**Diphtheria Immunization Coverage (`Diphtheria`)**: This variable reflects the percentage of one-year-olds who have received the DTP3 vaccine (diphtheria, tetanus toxoid, and pertussis).
+Immunization data is typically gathered through health system reports or household surveys and is recorded as a numeric percentage.
+
+**Schooling (`Schooling`)**: This measures the average number of years of schooling for individuals in the population.
+Data on schooling is sourced from education ministry reports or international databases like UNESCO, and it is recorded in years as a numeric value.
+
+**Medical Doctors (`pct_doctor`)**: This variable represents the number of medical doctors per 10,000 people in the population.
+It is typically reported by national health ministries or equivalent institutions.
+The dataset standardizes this figure as a numeric value to facilitate cross-country comparisons.
+
+**Nursing and Midwifery Personnel (`pct_nursing`)**: This measures the number of nursing and midwifery personnel per 10,000 people in the population.
+It is collected through similar national reporting systems and is recorded as a numeric value.
+
+## Data Cleaning
+
+The raw Life Expectancy dataset, along with two supplementary datasets for medical doctors and nursing personnel, underwent several cleaning steps to ensure it was accurate, consistent, and ready for analysis.
+Initially, the columns in the supplementary datasets were renamed for clarity and compatibility with the main dataset.
+A left join was then performed to merge information on `pct_doctor` and `pct_nursing` into the main dataset.
+Subsequently, the data was filtered to include only entries from 2004 to 2015, and rows with missing values were removed to maintain data integrity.
+Column names were standardized, and data types were adjusted as necessary to ensure compatibility with analytical processes.
+The cleaned dataset was then saved as a Parquet file for efficient storage and further analysis.
+
+## Outcome Variables
+The outcome variable is Life Expectancy.
+It represents the average number of years a person is expected to live, based on current health conditions and mortality rates.
+The model aims to understand how various factors, such as healthcare availability, education, and economic status, influence life expectancy across different countries.
+@fig-response displays the historical data of life expectancy across 195 countries from 2004 to 2015, comparing developed and developing countries.
+The data reveals distinct trends in life expectancy, with developed countries generally experiencing higher life expectancy compared to their developing counterparts.
 
 ```{r}
-#| label: fig-predictor
-#| echo: false
-#| warning: false
-
-ggplot(data, aes(x = LifeExpectancy)) +
-  geom_histogram(binwidth = 1, fill = "purple", color = "brown", alpha = 0.7) +
-  facet_wrap(~ Status) +
-  labs(
-    title = "Histogram of Life Expectancy by Status",
-    x = "Life Expectancy",
-    y = "Frequency"
-  ) +
-  theme_minimal()
-
-```
-
-Talk more about it.
-
-And also planes (@fig-planes).
-(You can change the height and width, but don't worry about doing that until you have finished every other aspect of the paper - Quarto will try to make it look nice and the defaults usually work well once you have enough text.)
-
-```{r}
-#| label: fig-planes
-#| fig-cap: Relationship between wing length and width
+#| label: fig-response
+#| fig-cap: "Distribution of life expectancy across different Status groups (Developed vs.
+Developing), highlighting the distinct patterns in their density.
+The mode of the density for life expectancy in developed countries tends to be around 81 years with a narrower variance, whereas in developing countries, the mode is approximately 74 years with a wider variance, reflecting disparities in health outcomes."
 #| echo: false
@@
 
+ggplot(data, aes(x = LifeExpectancy)) +
+  geom_histogram(aes(y = ..density..), 
+                 bins = 70, fill = "brown", color = "white", alpha = 0.7) +
+  geom_density(color = "royalblue", fill = "lightblue", alpha = 0.6) +
+  labs(
+    title = "Distribution of Life Expectancy by Status",
+    x = "Life Expectancy", y = "Density"
+  ) +
+  theme_minimal() +
+  facet_wrap(~Status)
+
 ```
 
-Talk way more about it.
+## Predictor Variables
 
-## Predictor variables
+The predictor variables (or independent variables) are the factors believed to influence life expectancy:
 
-Add graphs, tables and text.
+1.
+**Status (`Status`)**: This variable categorizes countries into "Developed" and "Developing" groups.
+The development status of a country plays a significant role in determining life expectancy, with developed countries generally having higher life expectancies due to better healthcare, sanitation, and infrastructure, as depicted in @fig-status.
 
-Use sub-sub-headings for each outcome variable and feel free to combine a few into one if they go together naturally.
+2.
+**Body Mass Index (`BMI`)**: This variable represents the average Body Mass Index of the entire population in a country.
+BMI is a widely used health indicator that reflects the prevalence of obesity or underweight populations.
+Extreme values of BMI can negatively impact life expectancy by contributing to chronic diseases such as diabetes and cardiovascular conditions.
+The BMI condition is depicted in @fig-bmi.
 
-## Cleaned Data
+3.
+**Diphtheria Immunization Coverage (`Diphtheria`)**: This variable represents the percentage of one-year-olds who have received the Diphtheria, Tetanus, and Pertussis (DTP3) immunization.
+Immunization coverage(@fig-diphtheria) is an important health indicator, as it reflects the effectiveness of a country's healthcare system in preventing infectious diseases, which in turn affects life expectancy.
+
+4.
+**Schooling (`Schooling`)**: The number of years of schooling in a country is used as a proxy for education levels.
+Higher education levels typically correlate with better health outcomes and, consequently, higher life expectancy, as educated individuals are more likely to engage in healthier behaviors and have access to better healthcare.
+The Educated condition is shown in @fig-schooling.
+
+5.
+**Medical Doctors (`pct_doctor`)**: This variable indicates the number of medical doctors per 10,000 people (@fig-doctor).
+Access to healthcare professionals is a key determinant of health outcomes, and a higher number of doctors generally corresponds with better healthcare services and longer life expectancy.
+
+6.
+**Nursing and Midwifery Personnel (`pct_nursing`)**: Similar to the number of doctors, this variable represents the number of nursing and midwifery personnel per 10,000 population (@fig-nursing).
+Nurses play a crucial role in delivering healthcare, and higher availability of nursing staff can contribute to improved health outcomes and increased life expectancy.
+
+7.
+**Health Expenditure (`TotalExpenditure`)**: This variable reflects the general government expenditure on health as a percentage of total government expenditure, as shown in @fig-totalExpenditure.
+Higher health spending is often associated with better healthcare infrastructure, which can lead to longer life expectancy due to better disease prevention and treatment options.
+
 ```{r}
-#Overview, delete later
-# Randomly select 10 rows from the data
-sampled_data <- data |> sample_n(10)
+#| label: fig-status
+#| fig-cap: "Histogram of Status, showing the counts of developed and developing countries in the dataset. There are 305 data points from developed countries and 788 from developing countries."
+#| echo: false
+#| warning: false
+#| message: false
 
-# Display the selected data
-sampled_data
+ggplot(data, aes(x = Status, fill = Status)) +
+  geom_bar(stat = "count") +
+  geom_text(stat = "count", aes(label = ..count..), vjust = 1.5, size = 5) +  # Move text to the bottom
+  labs(x = "Country Status", 
+       y = "Count", 
+       title = "Distribution of Countries by Status (Developed vs Developing)") +
+  scale_fill_manual(values = c("royalblue", "orange")) +
+  theme_minimal() +
+  theme(legend.position = "none")
 ```
 
-# Model
+```{r}
+#| label: fig-bmi
+#| fig-cap: "Density of BMI by Status, showing a higher concentration of BMI in developed regions around 24, suggesting a potential focus on obesity if the bin width were wider.
+The distribution for developing regions is fairly evenly distributed, with less concentration in any specific BMI range."
+#| echo: false
+#| warning: false
+#| message: false
+
+data_filtered <- data %>%
+  filter(BMI >= 18.5 & BMI <= 24.9)
+ggplot(data_filtered, aes(x = BMI, fill = Status)) +
+  geom_density(alpha = 0.5) +
+  labs(
+    title = "BMI Condition by Status (Normal Range)",
+    x = "BMI",
+    y = "Density"
+  ) +
+  theme_minimal() +
+  scale_fill_manual(values = c("royalblue", "red")) +  # Customize colors for Developed and Developing
+  facet_wrap(~Status)
+
+```
+
+```{r}
+#| label: fig-diphtheria
+#| fig-cap: "The plot shows similar immunization coverage conditions across developed and developing countries, with a noticeable increase in variance for developing countries due to some countries exhibiting low diphtheria immunization rates."
+#| echo: false
+#| warning: false
+#| message: false
+
+# Calculate variance for each Status group
+variance_data <- data %>%
+  group_by(Status) %>%
+  summarise(variance = var(Diphtheria, na.rm = TRUE))
+
+# Plot
+ggplot(data, aes(x = Diphtheria, y = Status, fill = Status)) +
+  geom_density_ridges(alpha = 0.7) +
+  # Add text for Variance on the left side
+  geom_text(data = variance_data, 
+            aes(x = 5, y = Status, label = paste("Variance:", round(variance, 1))),
+            size = 3, 
+            hjust = 0, 
+            color = "brown") +
+  labs(x = "Diphtheria Immunization Coverage (%)", 
+       y = "Country Status", 
+       title = "Distribution of Diphtheria Immunization by Country Status") +
+  scale_fill_manual(values = c("royalblue", "darkorange")) +
+  scale_x_continuous(limits = c(0, 100)) +  # Ensure x-axis is between 0 and 100
+  theme_minimal() +
+  theme(legend.position = "none")
+```
+
+```{r}
+#| label: fig-schooling
+#| fig-cap: "Violin plot with boxplot for Schooling by Status, highlighting a narrow distribution for developed regions with a higher median of 16 years of schooling.
+In contrast, developing countries show a wider variance and a relatively lower median of around 12.5 years, indicating significant disparities in educational access and attainment, with developed countries having more consistent and higher levels of education compared to the varied and generally lower levels in developing countries."
+#| echo: false
+#| warning: false
+#| message: false
+
+ggplot(data, aes(
+  x = Status,
+  y = Schooling,
+  fill = Status
+)) +
+  geom_violin(alpha = 0.5, position = position_dodge(width = 1)) + 
+  geom_boxplot(width = 0.1, color = "white", outlier.shape = NA, position = position_dodge(width = 1)) +
+  labs(
+    x = "Status",
+    y = "Schooling Years",
+    title = "Distribution of Schooling Years by Status"
+  ) +
+  scale_fill_manual(values = c("Developed" = "brown", "Developing" = "darkgreen")) + 
+  # Separate panels for each status
+  facet_wrap(~Status, scales = "free_x") +
+  theme_minimal() +
+  theme(axis.text.x = element_text(angle = 45, hjust = 1))
+
+```
+
+```{r}
+#| label: fig-totalExpenditure
+#| fig-cap: "Boxplot for government expenditure on health as a percentage of total expenditure, showing relatively high percentages for developed countries compared to developing countries, highlighting that developed countries' governments typically allocate more resources to health, though there are exceptions."
+#| echo: false
+#| warning: false
+#| message: false
+
+ggplot(data, aes(x = Status, y = TotalExpenditure, fill = Status)) +
+  # Adding boxplots on top of violin plots for more detail on median and quartiles
+  geom_boxplot(width = 0.5, alpha = 0.7, position = position_dodge(0.9)) +
+
+  labs(
+    title = "Total Government Health Expenditure by Status",
+    x = "Country Status",
+    y = "Total Government Expenditure on Health (%)"
+  ) +
+  theme_minimal() +
+  scale_fill_manual(values = c("skyblue", "orange")) +  # Customize colors for Developed and Developing
+  theme(legend.position = "none")  # Hide the legend for clarity
+
+```
+
+```{r}
+#| label: fig-doctor
+#| fig-cap: "Density plot for doctors per 10,000 population by Status, showing that the density of medical doctors in developed regions is relatively higher compared to developing regions.
+In developing countries, the density is concentrated at lower levels, suggesting that the healthcare system is generally stronger and more accessible in developed countries."
+#| echo: false
+#| warning: false
+#| message: false
+
+ggplot(data, aes(x = pct_doctor, y = Status, fill = Status)) +
+  geom_density_ridges(alpha = 0.7) +
+  labs(x = "Density of Medical Doctor", y = "Country Status", title = "Medical Doctor Condition") +
+  scale_fill_manual(values = c("royalblue", "darkorange")) +
+  theme_minimal() +
+  theme(legend.position = "none")
+
+```
+
+```{r}
+#| label: fig-nursing
+#| fig-cap: "Number of Nurses by Status with Highlighted Mean and Dotplot.
+The mean for developed countries is around 80, while the mean for developing countries is around 35, with many data points showing low nurse density in developing countries.
+Both status groups have medians lower than their means, suggesting a right-skewed distribution with a concentration of countries with fewer nurses."
+#| echo: false
+#| warning: false
+#| message: false
+
+ggplot(data, aes(x = Status, y = pct_nursing, fill = Status)) +
+  geom_boxplot(alpha = 0.8, outlier.shape = NA) +  # Boxplot for the distribution
+  geom_dotplot(binaxis = "y", stackdir = "center", dotsize = 0.5, alpha = 0.4, position = "dodge") +  # Smaller dots with dodge position
+  stat_summary(fun = "mean", geom = "point", shape = 18, size = 4, color = "black", fill = "yellow") +  # Highlight the mean
+  labs(
+    title = "Number of Nurses by Status with Highlighted Mean and Dotplot",
+    x = "Status",
+    y = "Number of Nurses per 10000"
+  ) +
+  theme_minimal() +
+  scale_fill_manual(values = c("brown", "darkgreen")) +  # Customize fill colors
+  theme(axis.text.x = element_text(angle = 45, hjust = 1))  # Tilt x-axis labels
+
+```
+\newpage
+
+# Model {#sec-model}
 
@@
 
-Define $y_i$ as the number of seconds that the plane remained aloft. Then $\beta_i$ is the wing width and $\gamma_i$ is the wing length, both measured in millimeters.
+Define $y_i$ as the Life Expectancy for country $i$. The factors $x_1$, $x_2$,...
+represent socioeconomic predictors such as BMI, Diphtheria, Schooling, TotalExpenditure, pct_doctor, and pct_nursing.
 
-\begin{align} 
-y_i|\mu_i, \sigma &\sim \mbox{Normal}(\mu_i, \sigma) \\
-\mu_i &= \alpha + \beta_i + \gamma_i\\
-\alpha &\sim \mbox{Normal}(0, 2.5) \\
-\beta &\sim \mbox{Normal}(0, 2.5) \\
-\gamma &\sim \mbox{Normal}(0, 2.5) \\
-\sigma &\sim \mbox{Exponential}(1)
-\end{align}
+```{=tex}
+\begin{align*} 
+\text{LifeExpectancy}_{\text{developed}} = &\ \beta_0 + \beta_1 \text{Diphtheria Immunization} + \beta_2 \text{Schooling} + \\
+&\ \beta_3 \text{Health Expenditure} + \beta_4 \text{Number of Doctors per 10000} + \epsilon_i
+\end{align*}
 
-We run the model in R [@citeR] using the `rstanarm` package of @rstanarm. We use the default priors from `rstanarm`.
+\begin{align*} 
+\text{LifeExpectancy}_{\text{developing}} = &\ \beta_0 + \beta_1 \text{BMI} + \beta_2 \text{Diphtheria Immunization} + \\
+&\ \beta_3 \text{Schooling} + \beta_4 \text{Number of Doctors per 10000} + \\
+&\ \beta_5 \text{Number of Nurses per 10000} + \epsilon_i
+\end{align*}
+```
 
-### Model justification
+## Model Justification
 
-We expect a positive relationship between the size of the wings and time spent aloft. In particular...
+The models were built using linear regression (LM), which is appropriate for predicting life expectancy as a continuous variable.
+This method captures the relationship between predictors and the response variable with interpretable coefficients.
+For the Developed model, predictors such as Diphtheria Immunization, Schooling, Total Expenditure, and the number of Doctors per 10,000 were selected as they are widely recognized to influence life expectancy in high-income countries.
+In the Developing model, we added BMI, Schooling, and both Doctors and Nurses per 10,000, reflecting healthcare access and lifestyle factors relevant to developing nations.
+Linear regression was chosen because it is effective for normally distributed continuous outcomes, which aligns with our dataset.
 
-We can use maths by including latex between dollar signs, for instance $\theta$.
+## Model Weaknesses and Limitations
 
-# Results
+Linear regression assumes that the residuals errors of the model are normally distributed.
+If this assumption is violated, it could lead to biased estimates and incorrect conclusions.
+This issue is especially important in cross-country data, where life expectancy may be influenced by diverse factors that do not follow a normal distribution.
 
-Our results are summarized in @tbl-modelresults.
+Predictors such as the number of doctors per 10,000 and the number of nurses per 10,000 might be correlated with each other.
+High correlation between predictors can cause multicollinearity, making it difficult to assess the individual contribution of each predictor.
 
+## Model Validation
+
+The models for both Developed and Developing countries were implemented using R, with applying out-of-sample testing, the data was randomly split into training and test sets, with 80% used for training the model and 20% reserved for testing.
+This allows us to assess how well the model performs on data that was not used during training.
+The Root Mean Square Error (RMSE) was calculated for each model to evaluate the performance and generalization to unseen data.
+A lower RMSE indicates better predictive performance.
+
+In @tbl-keyestimates, the Developed model is more efficient and better fitted primarily due to its higher R² (64.5% vs 51.5%), indicating that it explains a greater proportion of variance in life expectancy.
+Additionally, its RMSE value is lower (3.78 vs 5.06), showing more precise predictions.
+The Developing model has a larger sample size and variability, which might have diluted the effect of individual predictors.
+The better fit of the Developed model is partly due to the smaller, more homogeneous sample, allowing it to provide more reliable estimates with fewer outliers.
+
+### Model for Developed Countries
+
+For the Developed model, the predictors considered included Diphtheria Immunization, Schooling, Total Expenditure, and Number of Doctors per 10000.
+The coefficients and standard errors for these predictors were as follows:
+
+**Diphtheria Immunization**: Coefficient = 0.022, Standard Error = 0.025
+
+**Schooling**: Coefficient = 0.801, Standard Error = 0.151
+
+**Health Expenditure**: Coefficient = 0.056, Standard Error = 0.084
+
+**Number of Medical Doctors per 10000**: Coefficient = 0.018, Standard Error = 0.034
+
+These coefficients indicate that `Schooling` has the highest contribution to explaining life expectancy, followed by `Diphtheria Immunization`.
+The relatively small contribution of `#Doctors per 10000` suggests that doctor practices may have a less direct impact on life expectancy in developed countries, but this could be nuanced by other health factors not captured in the model.
+
+### Model for Developing Countries
+
+For the Developing model, the same predictors were considered, but their relationships with life expectancy differed. The coefficients and standard errors for each predictor were as follows:
+
+**BMI**: Coefficient = 0.040, Standard Error = 0.012
+
+**Diphtheria Immunization**: Coefficient = 0.036, Standard Error = 0.010
+
+**Schooling**: Coefficient = 1.727, Standard Error = 0.113
+
+**Number of Medical Doctors per 10000**: Coefficient = 0.090, Standard Error = 0.021
+
+**Number of Nurses per 10000**: Coefficient = -0.003, Standard Error = 0.003
+
+In the Developing model, `Schooling` again has the most significant impact, and `#Doctors per 10000` is also a substantial predictor.
+This highlights the importance of healthcare workforce availability in improving life expectancy in developing countries, unlike the developed context where Schooling was more prominent.
+The negative relationship of `Total Expenditure` suggests that in developing countries, spending in certain healthcare sectors could be misallocated or inefficient.
 ```{r}
+#| label: tbl-keyestimates
+#| tbl-cap: "Summary of key model estimates for Developed and Developing counties, including coefficients for predictors like BMI, Diphtheria, Schooling, and TotalExpenditure, pct_doctor, pct_nursing, with standard errors for each estimate.
+Model performance statistics, such as sample size, R², and adjusted R², are also displayed."
 #| echo: false
@@
 
-library(rstanarm)
+# load data and model
+lm_developed <- readRDS(here::here("models/lm_developed.rds"))
+lm_developing <- readRDS(here::here("models/lm_developing.rds"))
 
-first_model <-
-  readRDS(file = here::here("models/first_model.rds"))
+model_list <- list("Model for Developed" = lm_developed, "Model for Developing" = lm_developing)
+modelsummary(model_list, 
+             output = "kableExtra", 
+             statistic = "std.error", 
+             coef_map = c("BMI" = "BMI", 
+                          "Diphtheria" = "Diphtheria Immunization",
+                          "Schooling" = "Schooling",
+                          "TotalExpenditure" = "Health Expenditure",
+                          "pct_doctor" = "Number of Medical Doctors per 10000",
+                          "pct_nursing" = "Number of Nurses per 10000")) %>%
+  kable_styling(latex_options = c("striped", "scale_down"))
+```
+
+\newpage
+
+# Results {#sec-results}
+
+```{r}
+#| echo: false
+#| label: tbl-modelresults
+#| tbl-cap: "The prediction table indicates that the average life expectancy in developed countries is 77.23 years, while in developing countries, it is slightly lower at 72.1 years.
+This difference of approximately 5.1 years in life expectancy suggests a significant health and infrastructure disparity between developed and developing countries.
+These results align with broader global trends where developed countries typically benefit from better healthcare systems, higher standards of living, and advanced medical technology, contributing to longer life expectancies.
+"
+#| eval: true
+#| warning: false
+#| message: false
+
+# load data and model
+data <- read.csv(here::here("data/02-analysis_data/analysis_data.csv"))
+lm_developed <- readRDS(here::here("models/lm_developed.rds"))
+lm_developing <- readRDS(here::here("models/lm_developing.rds"))
+
+# make predictions
+data <- data %>%
+  mutate(
+    PredictedDeveloped = predict(lm_developed, newdata = data),
+    PredictedDeveloping = predict(lm_developing, newdata = data)
+  )
+
+# median and average predictions
+results <- data %>%
+  summarise(
+    AverageDeveloped = mean(PredictedDeveloped, na.rm = TRUE),
+    AverageDeveloping = mean(PredictedDeveloping, na.rm = TRUE)
+  )
+
+# display avg percent and normalized percent
+results %>%
+  select(AverageDeveloped, AverageDeveloping) %>%
+  kable(
+    col.names = c("Average Life Expectancy in Developed Countries", "Average Life Expectancy in Developing Countries"),
+    digits = 2,
+    booktabs = TRUE
+  ) %>%
+  kable_styling(
+    latex_options = c("striped", "scale_down")
+  )
 ```
@@
 #| eval: true
-#| label: tbl-modelresults
-#| tbl-cap: "Explanatory models of flight time based on wing width and wing length"
+#| label: fig-resultsplot
+#| fig-cap: "The prediction plots demonstrate that life expectancy in developed countries tends to remain higher with less variance, indicating stability in health outcomes due to better healthcare systems, more consistent access to medical services, and overall economic stability with less variance.
+In contrast, the life expectancy in developing countries exhibits a broader range of variation, with a more rapid growth trajectory.
+This suggests that while life expectancy is increasing in these countries, it is influenced by a variety of factors such as improvements in healthcare, economic development, and access to resources, all of which vary more significantly across different regions."
 #| warning: false
+#| message: false
 
-modelsummary::modelsummary(
-  list(
-    "First model" = first_model
-  ),
-  statistic = "mad",
-  fmt = 2
+data <- read.csv(here::here("data/02-analysis_data/analysis_data.csv"))
+lm_developed <- readRDS(here::here("models/lm_developed.rds"))
+lm_developing <- readRDS(here::here("models/lm_developing.rds"))
+
+# Make predictions
+data <- data %>%
+  mutate(
+    PredictedDeveloped = predict(lm_developed, newdata = data),
+    PredictedDeveloping = predict(lm_developing, newdata = data)
   )
+
+# Reshape data for plotting
+plot_data <- data %>%
+  select(Year, PredictedDeveloped, PredictedDeveloping) %>%
+  pivot_longer(
+    cols = c(PredictedDeveloped, PredictedDeveloping),
+    names_to = "Status",
+    values_to = "PredictedLifeExpectancy"
+  ) %>%
+  mutate(Status = case_when(
+    Status == "PredictedDeveloped" ~ "Developed",
+    Status == "PredictedDeveloping" ~ "Developing"
+  ))
+
+ggplot(plot_data, aes(x = Year, y = PredictedLifeExpectancy, color = Status)) +
+  geom_smooth(method = "loess", se = TRUE, size = 1) +
+  labs(
+    title = "Predicted Life Expectancy Over Time",
+    x = "Year",
+    y = "Predicted Life Expectancy (Years)",
+    color = "Status"
+  ) +
+  scale_color_manual(values = c("Developed" = "brown", "Developing" = "darkgreen")) +
+  theme_minimal() +
+  theme(
+    text = element_text(size = 12),
+    legend.position = "top"
+  )
+
 ```
+\newpage
 
-# Discussion
+# Discussion {#sec-discussion}
 
@@
 
-In @fig-ppcheckandposteriorvsprior-1 we implement a posterior predictive check. This shows...
-
-In @fig-ppcheckandposteriorvsprior-2 we compare the posterior with the prior. This shows...
-
-```{r}
-#| eval: true
-#| echo: false
-#| message: false
-#| warning: false
-#| label: fig-ppcheckandposteriorvsprior
-#| layout-ncol: 2
-#| fig-cap: "Examining how the model fits, and is affected by, the data"
-#| fig-subcap: ["Posterior prediction check", "Comparing the posterior with the prior"]
-
-pp_check(first_model) +
-  theme_classic() +
-  theme(legend.position = "bottom")
-
-posterior_vs_prior(first_model) +
-  theme_minimal() +
-  scale_color_brewer(palette = "Set1") +
-  theme(legend.position = "bottom") +
-  coord_flip()
-```
-
 ## Diagnostics
-
-@fig-stanareyouokay-1 is a trace plot. It shows... This suggests...
-
-@fig-stanareyouokay-2 is a Rhat plot. It shows... This suggests...
-
-```{r}
-#| echo: false
-#| eval: true
-#| message: false
-#| warning: false
-#| label: fig-stanareyouokay
-#| fig-cap: "Checking the convergence of the MCMC algorithm"
-#| fig-subcap: ["Trace plot", "Rhat plot"]
-#| layout-ncol: 2
-
-plot(first_model, "trace")
-
-plot(first_model, "rhat")
-```
 

### 2024-11-27 scripts/05-exploratory_data_analysis.R
Update 05-exploratory_data_analysis.R

@@ -1,39 +1,42 @@
 #### Preamble ####
-# Purpose: Perform exploratory data analysis on cleaned data
+# Purpose: This script performs exploratory data analysis (EDA) on the cleaned life expectancy data
 # Author: Manjun Zhu
-# Date: 23 November 2024
+# Date: 26 November 2024
 # Contact: karmen.zhu@mail.utoronto.ca
 # License: MIT
-# Pre-requisites: The `tidyverse`, 'janitor', 'here', and 'reshape2' packages
-# must be installed
+# Pre-requisites: The 'tidyverse', 'here', and 'lubridate' packages must be installed
 
 
 # load libraries
 library(tidyverse)
-library(janitor)
 library(here)
-library(reshape2)
-library(arrow)
+library(lubridate)
 
 # read cleaned data
-data <- read_parquet(here::here("data/02-analysis_data/analysis_data.parquet")) |>
-  clean_names()
+data <- read_parquet(here::here("data/02-analysis_data/analysis_data.parquet"))
 
-# quick overview of data structure
-glimpse(data)
+# Summary of numerical variables
+summary_stats <- data %>%
+  summarise(
+    LifeExpectancy = mean(LifeExpectancy, na.rm = TRUE),
+    Diphtheria = mean(Diphtheria, na.rm = TRUE),
+    TotalExpenditure = mean(TotalExpenditure, na.rm = TRUE),
+    BMI = mean(BMI, na.rm = TRUE),
+    Status = mean(as.numeric(Status), na.rm = TRUE),
+    Schooling = mean(Schooling, na.rm = TRUE),
+    pct_doctor = mean(pct_doctor, na.rm = TRUE),
+    pct_nursing = mean(pct_nursing, na.rm = TRUE),
+  )
 
-# print first few rows of data
-head(data)
+print(summary_stats)
 
-# print last few rows of data
-tail(data)
 
-# randomly sample 6 rows
-data |>
-  slice_sample(n = 6)
+# Status breakdown
+status_distribution <- data %>%
+  count(Status) %>%
+  mutate(percent = n / sum(n) * 100)
 
-# summary of the dataset
-summary(data)
+print(status_distribution)
 
 # number of NA values in each column
 missing_data <- data |>

### 2024-11-27 scripts/07-replications.R
Update 07-replications.R

@@ -1,19 +1,52 @@
 #### Preamble ####
-# Purpose: Replicated graphs from... [...UPDATE THIS...]
-# Author: Rohan Alexander [...UPDATE THIS...]
-# Date: 11 February 2023 [...UPDATE THIS...]
-# Contact: rohan.alexander@utoronto.ca [...UPDATE THIS...]
+# Purpose: This script performs out-of-sample testing and calculates the RMSE for the linear model trained in the previous script.
+# Author: Manjun Zhu
+# Date: 26 November 2024
+# Contact: karmen.zhu@mail.utoronto.ca
 # License: MIT
-# Pre-requisites: [...UPDATE THIS...]
-# Any other information needed? [...UPDATE THIS...]
+# Pre-requisites: The `tidyverse`, `caret`, and 'here' packages must be installed
 
 
 #### Workspace setup ####
 library(tidyverse)
-# [...UPDATE THIS...]
+library(caret)
+library(here)
 
-#### Load data ####
-# [...ADD CODE HERE...]
+# Load the cleaned dataset (adjust the file path as needed)
+data <- read_parquet(here::here("data/02-analysis_data/analysis_data.parquet"))
 
+# Clean the data again to ensure consistency with the model
+data_cleaned <- data %>%
+  filter(!is.na(LifeExpectancy), !is.na(Diphtheria), !is.na(Schooling), 
+         !is.na(TotalExpenditure), !is.na(pct_doctor)) %>%
+  mutate(
+    LifeExpectancy = as.numeric(LifeExpectancy),
+    Diphtheria = as.numeric(Diphtheria),
+    Schooling = as.integer(Schooling),
+    TotalExpenditure = as.integer(TotalExpenditure),
+    pct_doctor = as.integer(pct_doctor)
+  )
 
+# Split the data into training (80%) and testing (20%) sets
+set.seed(955) # For reproducibility
+Index <- createDataPartition(data_cleaned$LifeExpectancy,
+                                  p = 0.8,
+                                  list = FALSE,
+                                  times = 1
+)
+
+train_data <- data_cleaned[Index, ]
+test_data <- data_cleaned[-Index, ]
+
+# Load the saved model
+finisher_model <- readRDS(here::here("models/lm_developed.rds"))
+
+# Perform out-of-sample prediction on the test set
+predicted <- predict(finisher_model, newdata = test_data)
+
[…]

### 2024-11-27 scripts/07-replications.R
Update 07-replications.R

@@ -6,7 +6,6 @@
 # License: MIT
 # Pre-requisites: The `tidyverse`, `caret`, and 'here' packages must be installed
 
-
 #### Workspace setup ####
 library(tidyverse)
 library(caret)
@@ -49,4 +48,3 @@ rmse <- sqrt(mean((predicted - test_data$LifeExpectancy)^2))
 
 # Print RMSE value
 cat("Root Mean Squared Error (RMSE) on the test data: ", round(rmse, 2), "\n")
-

### 2024-11-27 scripts/03-clean_data.R
Update 03-clean_data.R

@@ -16,7 +16,7 @@ data <- read_csv("data/01-raw_data/raw_data.csv")
 analysis_data <- data %>%
   # Select the desired columns
   select(Country, Year, Status, Life.expectancy, Diphtheria, 
-         Total.expenditure, BMI, Schooling, pct_doctor, pct_nursing) %>%
+         Total.expenditure, Schooling, pct_doctor, pct_nursing) %>%
   # Filter rows where Year is between 2009 and 2015, and drop the unrealistic value
   filter(Year >= 2004 & Year <= 2015) %>%
   filter(pct_doctor >= 0 & pct_doctor <= 1500) %>%

### 2024-11-27 scripts/04-test_analysis_data.R
Update 04-test_analysis_data.R

@@ -53,7 +53,6 @@ test_results$na_values_test <- test_that("NA values", {
                info = "Schooling contains NA values.")
   expect_false(any(is.na(analysis_data$TotalExpenditure)), 
                info = "TotalExpenditure contains NA values.")
-  expect_false(any(is.na(analysis_data$BMI)), info = "BMI contains NA values.")
 })
 
 # Test if Year values are within expected range (2009-2015)
@@ -92,14 +91,6 @@ test_results$diphtheria_test <- test_that("Diphtheria vaccination rate", {
               info = "Diphtheria vaccination rates are out of expected range 
               (0% to 100%).")
 })
-
-# Test if BMI values are within expected range (e.g., 10 to 50)
-test_results$bmi_test <- test_that("BMI range", {
-  expect_true(all(analysis_data$BMI >= 10 & 
-                    analysis_data$BMI <= 50, na.rm = TRUE),
-              info = "BMI values are out of expected range (10 to 50).")
-})
-
 
 # Test if Schooling values are within expected range (3 to 30)
 test_results$schooling_range_test <- test_that("Schooling range", {

### 2024-11-27 scripts/05-exploratory_data_analysis.R
Update 05-exploratory_data_analysis.R

@@ -21,7 +21,6 @@ summary_stats <- data %>%
     LifeExpectancy = mean(LifeExpectancy, na.rm = TRUE),
     Diphtheria = mean(Diphtheria, na.rm = TRUE),
     TotalExpenditure = mean(TotalExpenditure, na.rm = TRUE),
-    BMI = mean(BMI, na.rm = TRUE),
     Status = mean(as.numeric(Status), na.rm = TRUE),
     Schooling = mean(Schooling, na.rm = TRUE),
     pct_doctor = mean(pct_doctor, na.rm = TRUE),

### 2024-11-27 scripts/06-model_data.R
Update 06-model_data.R

@@ -1,10 +1,10 @@
 #### Preamble ####
-# Purpose: Run a linear regression model to predict Life Expectancy between Developed and Developing Countries
+# Purpose: This script performs out-of-sample testing and calculates the RMSE for the linear model trained in the previous script.
 # Author: Manjun Zhu
-# Date: 24 November 2024
+# Date: 26 November 2024
 # Contact: karmen.zhu@mail.utoronto.ca
 # License: MIT
-# Pre-requisites: The `tidyverse` and 'here' packages must be installed
+# Pre-requisites: The `tidyverse`, `caret`, and 'here' packages must be installed
 
 
 #### Workspace setup ####
@@ -21,7 +21,6 @@ data <- data %>%
     LifeExpectancy = as.numeric(LifeExpectancy),
     Status = as.factor(Status),
     Diphtheria = as.numeric(Diphtheria),
-    BMI = as.numeric(BMI),
     TotalExpenditure = as.numeric(TotalExpenditure),
     Schooling = as.numeric(Schooling),
     pct_doctor = as.numeric(pct_doctor),
@@ -30,12 +29,12 @@ data <- data %>%
 
 ### Model data ####
 # linear model for Developed Countries' Life Expectancy based on socioeconomic factors
-lm_developed <- lm(LifeExpectancy ~  Diphtheria + Schooling + TotalExpenditure + pct_doctor, 
+lm_developed <- lm(LifeExpectancy ~  Diphtheria + Schooling + TotalExpenditure * pct_doctor * pct_nursing, 
                    data = data %>% filter(Status == "Developed"))
 
 
 # linear model for Developing Countries' Life Expectancy based on socioeconomic factors
-lm_developing <- lm(LifeExpectancy ~ BMI + Diphtheria + Schooling + pct_doctor + pct_nursing, 
+lm_developing <- lm(LifeExpectancy ~ Diphtheria + Schooling + TotalExpenditure * pct_doctor * pct_nursing, 
                     data = data %>% filter(Status == "Developing"))
 
 #### Save model ####

### 2024-11-27 paper/paper.qmd
Add files via upload

@@
 The response variable is Life Expectancy, while the predictors include Status (Developed vs.
-Developing), Diphtheria Immunization Coverage, Government Total Expenditure, Numbers of Doctors and Nurses per 10000, Schooling and BMI.
+Developing), Diphtheria Immunization Coverage, Government Total Expenditure, Numbers of Doctors and Nurses per 10000, and Schooling.
 The analysis seeks to quantify the influence of these variables, providing insights into the disparities in life expectancy and identifying factors that contribute to these differences.
@@
 Key variables include Life Expectancy, Status (Developed vs.
-Developing), Diphtheria Immunization Coverage, Government Health Expenditure as percentage of Total Expenditure, Numbers of Doctors and Nurses per 10000, Schooling and BMI.
+Developing), Diphtheria Immunization Coverage, Government Health Expenditure as percentage of Total Expenditure, Numbers of Doctors and Nurses per 10000, and Schooling.
 
@@
 **Status (`Status`)**: This categorical variable identifies whether a country is classified as "Developed" or "Developing." The classification is based on economic and social indicators defined by the World Health Organization (WHO) or related agencies.
-
-**Body Mass Index (`BMI`)**: This variable captures the average body mass index of the population, calculated as weight in kilograms divided by the square of height in meters.
-It is an aggregate measure derived from population health surveys and recorded as a numeric value.
 
@@
 2.
-**Body Mass Index (`BMI`)**: This variable represents the average Body Mass Index of the entire population in a country.
-BMI is a widely used health indicator that reflects the prevalence of obesity or underweight populations.
-Extreme values of BMI can negatively impact life expectancy by contributing to chronic diseases such as diabetes and cardiovascular conditions.
-The BMI condition is depicted in @fig-bmi.
-
-3.
 **Diphtheria Immunization Coverage (`Diphtheria`)**: This variable represents the percentage of one-year-olds who have received the Diphtheria, Tetanus, and Pertussis (DTP3) immunization.
@@
 
-4.
+3.
 **Schooling (`Schooling`)**: The number of years of schooling in a country is used as a proxy for education levels.
@@
 
-5.
+4.
 **Medical Doctors (`pct_doctor`)**: This variable indicates the number of medical doctors per 10,000 people (@fig-doctor).
@@
 
-6.
+5.
 **Nursing and Midwifery Personnel (`pct_nursing`)**: Similar to the number of doctors, this variable represents the number of nursing and midwifery personnel per 10,000 population (@fig-nursing).
@@
   theme(legend.position = "none")
-```
-
-```{r}
-#| label: fig-bmi
-#| fig-cap: "Density of BMI by Status, showing a higher concentration of BMI in developed regions around 24, suggesting a potential focus on obesity if the bin width were wider.
-The distribution for developing regions is fairly evenly distributed, with less concentration in any specific BMI range."
-#| echo: false
-#| warning: false
-#| message: false
-
-data_filtered <- data %>%
-  filter(BMI >= 18.5 & BMI <= 24.9)
-ggplot(data_filtered, aes(x = BMI, fill = Status)) +
-  geom_density(alpha = 0.5) +
-  labs(
-    title = "BMI Condition by Status (Normal Range)",
-    x = "BMI",
-    y = "Density"
-  ) +
-  theme_minimal() +
-  scale_fill_manual(values = c("royalblue", "red")) +  # Customize colors for Developed and Developing
-  facet_wrap(~Status)
-
 ```
@@
 
-The goal of our modelling strategy is twofold. Firstly,...
-
-Here we briefly describe the Bayesian analysis model used to investigate... Background details and diagnostics are included in [Appendix -@sec-model-details].
+The goal of this analysis is to construct a linear model to predict life expectancy (`LifeExpectancy`), which represents the average number of years an individual is expected to live.
+This model incorporates several predictor variables that capture economic and social factors influencing life expectancy.
 
@@
 
-Define $y_i$ as the Life Expectancy for country $i$. The factors $x_1$, $x_2$,...
-represent socioeconomic predictors such as BMI, Diphtheria, Schooling, TotalExpenditure, pct_doctor, and pct_nursing.
+The models assume a linear relationship between the predictors and life expectancy, where each predictor contributes independently or interactively to the variation in the outcome variable.
+Define $y_i$ as the Life Expectancy for country $i$.
+The factors $x_1$, $x_2$,...represent socioeconomic predictors such as Diphtheria, Schooling, TotalExpenditure, pct_doctor, pct_nursing, and interaction term.
+
+The linear regression model for developed and developing regions are specified as follows:
 
 ```{=tex}
+
 \begin{align*} 
 \text{LifeExpectancy}_{\text{developed}} = &\ \beta_0 + \beta_1 \text{Diphtheria Immunization} + \beta_2 \text{Schooling} + \\
-&\ \beta_3 \text{Health Expenditure} + \beta_4 \text{Number of Doctors per 10000} + \epsilon_i
+&\ \beta_3 \text{Health Expenditure} + \beta_4 \text{Number of Doctors per 10000} + \\
+&\ \beta_5 \text{Number of Nurses per 10000} + \\
+&\ \beta_6 \text{Health Expenditure} \times \text{Number of Doctors per 10000} \times \text{Number of Nurses per 10000} + \epsilon_i
 \end{align*}
@@
 \begin{align*} 
-\text{LifeExpectancy}_{\text{developing}} = &\ \beta_0 + \beta_1 \text{BMI} + \beta_2 \text{Diphtheria Immunization} + \\
-&\ \beta_3 \text{Schooling} + \beta_4 \text{Number of Doctors per 10000} + \\
-&\ \beta_5 \text{Number of Nurses per 10000} + \epsilon_i
+\text{LifeExpectancy}_{\text{developing}} = &\ \beta_0 + \beta_1 \text{Diphtheria Immunization} + \beta_2 \text{Schooling} + \\
+&\ \beta_3 \text{Health Expenditure} + \beta_4 \text{Number of Doctors per 10000} + \\
+&\ \beta_5 \text{Number of Nurses per 10000} + \\
+&\ \beta_6 \text{Health Expenditure} \times \text{Number of Doctors per 10000} \times \text{Number of Nurses per 10000} + \epsilon_i
 \end{align*}
+
 ```
+
+Where:
+
+-   `LifeExpectancy`: Represents the average number of years a person is expected to live in a specific country, based on current mortality trends.
+
+-   `Diphtheria Immunization`: The percentage of children who received the diphtheria-tetanus-pertussis (DTP3) vaccine, serving as a proxy for the quality and reach of immunization programs.
+
+-   `Schooling`: Average years of schooling for individuals aged 25 and above, reflecting the level of education in the population.
+
+-   `Health Expenditure`: The percentage of government expenditure allocated to healthcare, indicating the priority given to public health.
+
+-   `Number of Doctors per 10,000`: The availability of medical doctors per 10,000 people, capturing the accessibility of professional healthcare services.
+
+-   `Number of Nurses per 10,000`: The availability of nursing staff per 10,000 people, further representing healthcare capacity.
+
+-   Interaction Term: The combined effect of health expenditure, doctors, and nurses on life expectancy, considering how these variables may amplify or mitigate each other’s impact.
+
+-   epsilon is the error term, assumed to be normally distributed with a mean of 0 and constant variance $ϵ_i​∼N(0,\sigma^2)$
 
@@
 For the Developed model, predictors such as Diphtheria Immunization, Schooling, Total Expenditure, and the number of Doctors per 10,000 were selected as they are widely recognized to influence life expectancy in high-income countries.
-In the Developing model, we added BMI, Schooling, and both Doctors and Nurses per 10,000, reflecting healthcare access and lifestyle factors relevant to developing nations.
+In the Developing model, we added Schooling, and both Doctors and Nurses per 10,000, reflecting healthcare access and lifestyle factors relevant to developing nations.
 Linear regression was chosen because it is effective for normally distributed continuous outcomes, which aligns with our dataset.
@@
 
-**BMI**: Coefficient = 0.040, Standard Error = 0.012
-
 **Diphtheria Immunization**: Coefficient = 0.036, Standard Error = 0.010
@@
 #| label: tbl-keyestimates
-#| tbl-cap: "Summary of key model estimates for Developed and Developing counties, including coefficients for predictors like BMI, Diphtheria, Schooling, and TotalExpenditure, pct_doctor, pct_nursing, with standard errors for each estimate.
+#| tbl-cap: "Summary of key model estimates for Developed and Developing counties, including coefficients for predictors like Diphtheria, Schooling, and TotalExpenditure, pct_doctor, pct_nursing, with standard errors for each estimate.
 Model performance statistics, such as sample size, R², and adjusted R², are also displayed."
@@
              statistic = "std.error", 
-             coef_map = c("BMI" = "BMI", 
-                          "Diphtheria" = "Diphtheria Immunization",
+             coef_map = c("Diphtheria" = "Diphtheria Immunization",
                           "Schooling" = "Schooling",
@@
 #| label: tbl-modelresults
-#| tbl-cap: "The prediction table indicates that the average life expectancy in developed countries is 77.23 years, while in developing countries, it is slightly lower at 72.1 years.
+#| tbl-cap: "The prediction table indicates that the average life expectancy in developed countries is 78.53 years, while in developing countries, it is slightly lower at 71.68 years.
 This difference of approximately 5.1 years in life expectancy suggests a significant health and infrastructure disparity between developed and developing countries.

### 2024-11-27 paper/paper.qmd
Add files via upload

@@
 format: pdf
+toc: true
+toc-depth: 3
 number-sections: true
@@
 
+# load libraries
 library(dplyr)
@@
 
+# load data
 data <- read.csv(here::here("data/02-analysis_data/analysis_data.csv"))
@@
 
-This study aims to address these gaps by developing a linear model to analyze the impact of a nation’s status, healthcare factors, and economic indicators on life expectancy.
-The response variable is Life Expectancy, while the predictors include Status (Developed vs.
-Developing), Diphtheria Immunization Coverage, Government Total Expenditure, Numbers of Doctors and Nurses per 10000, and Schooling.
-The analysis seeks to quantify the influence of these variables, providing insights into the disparities in life expectancy and identifying factors that contribute to these differences.
+This study aims to examine the factors influencing life expectancy across different nations by developing a linear model that incorporates both healthcare and economic variables.
+The response variable in the analysis is life expectancy, while the predictors include national status (Developed vs.
+Developing), diphtheria immunization coverage, government healthcare expenditure as a percentage of total expenditure, and the number of doctors and nurses per 10,000 people.
+Additionally, schooling is considered as an important educational factor.
+By quantifying the effects of these predictors, the study seeks to shed light on disparities in life expectancy between developed and developing nations, and to identify key factors contributing to these differences.
+Through this analysis, we aim to contribute valuable insights for policy development focused on improving healthcare and economic outcomes.
 
@@
 ```
+\newpage
 
@@
 \begin{align*} 
-\text{LifeExpectancy}_{\text{developed}} = &\ \beta_0 + \beta_1 \text{Diphtheria Immunization} + \beta_2 \text{Schooling} + \\
-&\ \beta_3 \text{Health Expenditure} + \beta_4 \text{Number of Doctors per 10000} + \\
-&\ \beta_5 \text{Number of Nurses per 10000} + \\
-&\ \beta_6 \text{Health Expenditure} \times \text{Number of Doctors per 10000} \times \text{Number of Nurses per 10000} + \epsilon_i
+\text{LifeExpectancy}_{\text{developing}} = &\ \beta_0 + \beta_1 \text{Diphtheria Immunization}_i + \beta_2 \text{Schooling}_i + \\
+&\ \beta_3 \text{Health Expenditure}_i + \beta_4 \text{Number of Doctors per 10000}_i + \\
+&\ \beta_5 \text{Number of Nurses per 10000}_i + \\
+&\ \beta_6 \text{Health Expenditure}_i \times \text{Doctors per 10000}_i \times \text{Nurses per 10000}_i + \epsilon_i
 \end{align*}
-
 \begin{align*} 
-\text{LifeExpectancy}_{\text{developing}} = &\ \beta_0 + \beta_1 \text{Diphtheria Immunization} + \beta_2 \text{Schooling} + \\
-&\ \beta_3 \text{Health Expenditure} + \beta_4 \text{Number of Doctors per 10000} + \\
-&\ \beta_5 \text{Number of Nurses per 10000} + \\
-&\ \beta_6 \text{Health Expenditure} \times \text{Number of Doctors per 10000} \times \text{Number of Nurses per 10000} + \epsilon_i
+\text{LifeExpectancy}_{\text{developing}} = &\ \beta_0 + \beta_1 \text{Diphtheria Immunization}_i + \beta_2 \text{Schooling}_i + \\
+&\ \beta_3 \text{Health Expenditure}_i + \beta_4 \text{Number of Doctors per 10000}_i + \\
+&\ \beta_5 \text{Number of Nurses per 10000}_i + \\
+&\ \beta_6 \text{Health Expenditure}_i \times \text{Doctors per 10000}_i \times \text{Nurses per 10000}_i + \epsilon_i
 \end{align*}
@@
 
+The model assumes that the predictors have a linear relationship with the response variable, i.e., the life expectancy is linearly related to the selected features.
+We are interested in estimating the coefficients $\beta_1$, $\beta_2$,...,$\beta_4$ which represent the effect of each predictor and interaction term in the outcome.
+
 ## Model Justification
 
-The models were built using linear regression (LM), which is appropriate for predicting life expectancy as a continuous variable.
-This method captures the relationship between predictors and the response variable with interpretable coefficients.
-For the Developed model, predictors such as Diphtheria Immunization, Schooling, Total Expenditure, and the number of Doctors per 10,000 were selected as they are widely recognized to influence life expectancy in high-income countries.
-In the Developing model, we added Schooling, and both Doctors and Nurses per 10,000, reflecting healthcare access and lifestyle factors relevant to developing nations.
-Linear regression was chosen because it is effective for normally distributed continuous outcomes, which aligns with our dataset.
+The linear regression model was chosen for this analysis due to its simplicity and interpretability.
+The coefficients from this model directly show how each predictor affects life expectancy, assuming other variables remain constant.
+This makes it ideal for understanding the impact of healthcare and economic factors on life expectancy in both developed and developing countries.
+
+For the developed countries model, predictors like Diphtheria Immunization, Schooling, Total Expenditure, and the number of Doctors per 10,000 were included.
+These variables are well-established factors influencing life expectancy in high-income nations [@life_expectancy_who; @world_health_statistics].
+In the developing countries model, additional predictors such as Schooling, and healthcare access variables like Doctors and Nurses per 10,000 were added to account for differences in healthcare infrastructure.
+
+Linear regression is suitable for our dataset, as it assumes a normally distributed continuous response variable.
+While more complex models were considered, linear regression was preferred for its transparency and ease of interpretation, aligning with the goals of this analysis.
+Additionally, previous studies have shown that healthcare access, education, and economic factors play a significant role in determining life expectancy, particularly in developing countries [@Hanushek2016; @He2020].
+Including these predictors ensures the model is built on sound reasoning and established domain knowledge.
 

### 2024-11-27 scripts/00-simulate_data.R
Update 00-simulate_data.R

@@ -1,7 +1,7 @@
 #### Preamble ####
 # Purpose: Simulates a dataset of Life Expectancy
 # Author: Manjun Zhu
-# Date: 16 November 2024
+# Date: 27 November 2024
 # Contact: karmen.zhu@mail.utoronto.ca
 # License: MIT
 # Pre-requisites: The `tidyverse` package must be installed
@@ -13,50 +13,53 @@ library(tidyverse)
 set.seed(955)
 
 # Generate the number of rows for the dataset
-n <- 50
+n <- 500
 
 # Simulated data
 # Define countries and their status
-countries <- data.frame(
-  Country = c("USA", "Canada", "Germany", "China", "India",
-              "Brazil", "Nigeria", "South Africa"),
-  Status = c("Developed", "Developed", "Developed", "Developing",
-             "Developing", "Developing", "Developing", "Developing")
-)
+data <- function(n) {
+  country_names <- c("Australia", "Canada", "Germany", "Japan", "United States", 
+                     "Brazil", "India", "South Africa", "China", "Russia", 
+                     "France", "Italy", "United Kingdom", "South Korea", "Mexico", 
+                     "Argentina", "Spain", "Saudi Arabia", "Egypt", "Turkey", 
+                     "Nigeria", "Indonesia", "Malaysia", "Thailand", "Vietnam", 
+                     "Chile", "Poland", "Israel", "Pakistan", "Singapore", "Netherlands")
+  
+  # Randomly sample countries from the list
+  Country <- sample(country_names, n, replace = TRUE)
+  
+  # Randomly generate the data for each variable based on the summary statistics provided
+  Year <- sample(2004:2015, n, replace = TRUE)
+  
+  # Status (Developed/Developing) based on region or income classification
+  Status <- sample(c("Developed", "Developing"), n, replace = TRUE, prob = c(0.5, 0.5))
+  
+  # LifeExpectancy simulated around a mean of 72.24 with some variation
+  LifeExpectancy <- rnorm(n, mean = 72.24, sd = 10)
+  
+  # Diphtheria immunization rates between 4 and 99
+  Diphtheria <- runif(n, min = 4, max = 99)
+  
+  # TotalExpenditure between 1.12 and 14.39
+  TotalExpenditure <- runif(n, min = 1.12, max = 14.39)
+  
[…]

### 2024-11-27 scripts/00-simulate_data.R
Update 00-simulate_data.R

@@ -10,6 +10,10 @@
 
 #### Workspace setup ####
 library(tidyverse)
+library(here)
+library(lubridate)
+
+# set seed
 set.seed(955)
 
 # Generate the number of rows for the dataset
@@ -64,3 +68,4 @@ data <- data(500)
 
 #### Save data ####
 write_csv(data, "data/00-simulated_data/simulated_data.csv")
+write_parquet(data, "data/00-simulated_data/simulated_data.parquet")

### 2024-11-27 scripts/01-test_simulated_data.R
Update 01-test_simulated_data.R

@@ -1,7 +1,7 @@
 #### Preamble ####
 # Purpose: Tests the structure and validity of the simulated Life Expectancy
 # Author: Manjun Zhu
-# Date: 16 November 2024
+# Date: 27 November 2024
 # Contact: karmen.zhu@mail.utoronto.ca
 # License: MIT
 # Pre-requisites: 
@@ -12,8 +12,10 @@
 
 #### Workspace setup ####
 library(tidyverse)
+library(here)
 
-data <- read_csv("data/00-simulated_data/simulated_data.csv")
+data <- read_parquet(here::here("data/00-simulated_data/simulated_data.parquet"))
+
 
 # Test if the data was successfully loaded
 if (exists("data")) {
@@ -24,18 +26,18 @@ if (exists("data")) {
 
 #### Test data ####
 
-# Check if the dataset has the expected number of rows (e.g., 50 rows for the sample)
-if (nrow(data) == 50) {
-  message("Test Passed: The dataset has 50 rows.")
+# Check if the dataset has the expected number of rows (e.g., 500 rows for the full dataset)
+if (nrow(data) == 500) {
+  message("Test Passed: The dataset has 500 rows.")
 } else {
-  stop("Test Failed: The dataset does not have 50 rows.")
+  stop("Test Failed: The dataset does not have 500 rows.")
 }
 
-# Check if the dataset has 8 columns
-if (ncol(data) == 8) {
-  message("Test Passed: The dataset has 8 columns.")
+# Check if the dataset has 9 columns
+if (ncol(data) == 9) {
+  message("Test Passed: The dataset has 9 columns.")
 } else {
-  stop("Test Failed: The dataset does not have 7 columns.")
+  stop("Test Failed: The dataset does not have 9 columns.")
 }
 
 # Check if all countries are valid and non-empty
@@ -53,9 +55,9 @@ if (all(data$Status %in% valid_statuses)) {
   stop("Test Failed: The 'Status' column contains invalid entries.")
 }
 
-# Check if the 'Year' column includes only values between 2009 and 2015
-if (all(data$Year >= 2009 & data$Year <= 2015)) {
-  message("Test Passed: The 'Year' column contains only valid years (2009-2015).")
+# Check if the 'Year' column includes only values between 2004 and 2015
+if (all(data$Year >= 2004 & data$Year <= 2015)) {
+  message("Test Passed: The 'Year' column contains only valid years (2004-2015).")
 } else {
[…]

### 2024-11-27 scripts/00-simulate_data.R
Update 00-simulate_data.R

@@ -67,5 +67,4 @@ data <- data(500)
 
 
 #### Save data ####
-write_csv(data, "data/00-simulated_data/simulated_data.csv")
 write_parquet(data, "data/00-simulated_data/simulated_data.parquet")

### 2024-11-27 scripts/03-clean_data.R
Update 03-clean_data.R

@@ -30,5 +30,4 @@ colnames(analysis_data)[colnames(analysis_data) ==
                  "Total.expenditure"] <- "TotalExpenditure"
 
 #### Save data ####
-write_csv(analysis_data, "data/02-analysis_data/analysis_data.csv")
 write_parquet(analysis_data, "data/02-analysis_data/analysis_data.parquet")

### 2024-12-04 scripts/00-simulate_data.R
update scripts directory

@@ -1,70 +1,0 @@
-#### Preamble ####
-# Purpose: Simulates a dataset of Life Expectancy
-# Author: Manjun Zhu
-# Date: 27 November 2024
-# Contact: karmen.zhu@mail.utoronto.ca
-# License: MIT
-# Pre-requisites: The `tidyverse` package must be installed
-# Any other information needed? Make sure you are in the `starter_folder` rproj
-
-
-#### Workspace setup ####
-library(tidyverse)
-library(here)
-library(lubridate)
-
-# set seed
-set.seed(955)
-
-# Generate the number of rows for the dataset
-n <- 500
-
-# Simulated data
-# Define countries and their status
-data <- function(n) {
-  country_names <- c("Australia", "Canada", "Germany", "Japan", "United States", 
-                     "Brazil", "India", "South Africa", "China", "Russia", 
-                     "France", "Italy", "United Kingdom", "South Korea", "Mexico", 
-                     "Argentina", "Spain", "Saudi Arabia", "Egypt", "Turkey", 
-                     "Nigeria", "Indonesia", "Malaysia", "Thailand", "Vietnam", 
-                     "Chile", "Poland", "Israel", "Pakistan", "Singapore", "Netherlands")
-  
-  # Randomly sample countries from the list
-  Country <- sample(country_names, n, replace = TRUE)
-  
-  # Randomly generate the data for each variable based on the summary statistics provided
-  Year <- sample(2004:2015, n, replace = TRUE)
-  
-  # Status (Developed/Developing) based on region or income classification
-  Status <- sample(c("Developed", "Developing"), n, replace = TRUE, prob = c(0.5, 0.5))
-  
-  # LifeExpectancy simulated around a mean of 72.24 with some variation
-  LifeExpectancy <- rnorm(n, mean = 72.24, sd = 10)
-  
-  # Diphtheria immunization rates between 4 and 99
-  Diphtheria <- runif(n, min = 4, max = 99)
-  
-  # TotalExpenditure between 1.12 and 14.39
-  TotalExpenditure <- runif(n, min = 1.12, max = 14.39)
-  
-  # Schooling years between 3.1 and 20.4
-  Schooling <- runif(n, min = 3.1, max = 20.4)
-  
-  # pct_doctor between 0.13 and 75.67
-  pct_doctor <- runif(n, min = 0.13, max = 75.67)
-  
-  # pct_nursing between 0.58 and 174.30
[…]

### 2024-12-04 scripts/01-test_simulated_data.R

@@ -1,112 +1,0 @@
-#### Preamble ####
-# Purpose: Tests the structure and validity of the simulated Life Expectancy
-# Author: Manjun Zhu
-# Date: 27 November 2024
-# Contact: karmen.zhu@mail.utoronto.ca
-# License: MIT
-# Pre-requisites: 
-  # - The `tidyverse` package must be installed and loaded
-  # - 00-simulate_data.R must have been run
-# Any other information needed? Make sure you are in the `starter_folder` rproj
-
-
-#### Workspace setup ####
-library(tidyverse)
-library(here)
-
-data <- read_parquet(here::here("data/00-simulated_data/simulated_data.parquet"))
-
-
-# Test if the data was successfully loaded
-if (exists("data")) {
-  message("Test Passed: The dataset was successfully loaded.")
-} else {
-  stop("Test Failed: The dataset could not be loaded.")
-}
-
-#### Test data ####
-
-# Check if the dataset has the expected number of rows (e.g., 500 rows for the full dataset)
-if (nrow(data) == 500) {
-  message("Test Passed: The dataset has 500 rows.")
-} else {
-  stop("Test Failed: The dataset does not have 500 rows.")
-}
-
-# Check if the dataset has 9 columns
-if (ncol(data) == 9) {
-  message("Test Passed: The dataset has 9 columns.")
-} else {
-  stop("Test Failed: The dataset does not have 9 columns.")
-}
-
-# Check if all countries are valid and non-empty
-if (all(data$Country != "")) {
-  message("Test Passed: All countries have valid non-empty names.")
-} else {
-  stop("Test Failed: Some country names are missing or empty.")
-}
-
-# Check if the 'Status' column contains only "Developed" or "Developing"
-valid_statuses <- c("Developed", "Developing")
-if (all(data$Status %in% valid_statuses)) {
-  message("Test Passed: The 'Status' column contains only valid statuses.")
-} else {
-  stop("Test Failed: The 'Status' column contains invalid entries.")
-}
-
-# Check if the 'Year' column includes only values between 2004 and 2015
-if (all(data$Year >= 2004 & data$Year <= 2015)) {
-  message("Test Passed: The 'Year' column contains only valid years (2004-2015).")
-} else {
-  stop("Test Failed: The 'Year' column contains invalid years.")
-}
-
-# Check if there are no missing values in the dataset
-if (all(!is.na(data))) {
[…]

### 2024-12-04 scripts/02-download_data.R

@@ -1,42 +1,0 @@
-#### Preamble ####
-# Purpose: Downloads and saves the data from 'https://www.kaggle.com/datasets/kumarajarshi/life-expectancy-who'
-# Author: Manjun Zhu
-# Date: 24 November 2024
-# Contact: karmen.zhu@utoronto.ca
-# License: University of Toronto
-# Pre-requisites: No
-# Any other information needed? Thanks to Kaggle opendataset
-
-
-#### Workspace setup ####
-library(tidyverse)
-library(dplyr)
-
-#### Download data ####
-raw_data <- read.csv(file = "data/01-raw_data/LifeExpectancyData.csv")
-
-# Load the supplementary dataset
-doctor <- read.csv("data/01-raw_data/medicalDoctors.csv")
-nursing <- read.csv("data/01-raw_data/nursingAndMidwife.csv")
-
-# Ensure the column names align (modify as needed)
-# Example: rename 'Location' in doctor to 'Country'
-colnames(doctor)[colnames(doctor) == "Location"] <- "Country"
-colnames(doctor)[colnames(doctor) == "Period"] <- "Year"
-colnames(doctor)[colnames(doctor) == "First.Tooltip"] <- "pct_doctor"
-colnames(nursing)[colnames(nursing) == "Location"] <- "Country"
-colnames(nursing)[colnames(nursing) == "Period"] <- "Year"
-colnames(nursing)[colnames(nursing) == "First.Tooltip"] <- "pct_nursing"
-
-# Merge raw_data with doctor dataset
-test_data <- raw_data %>%
-  left_join(doctor %>% select(Country, Year, pct_doctor), by = c("Country", "Year"))
-
-# Merge test_data with nursing dataset
-raw_data <- test_data %>%
-  left_join(nursing %>% select(Country, Year, pct_nursing), by = c("Country", "Year"))
-
-
-#### Save data ####
-# change raw_data to whatever name you assigned when you downloaded it.
-write_csv(raw_data, "data/01-raw_data/raw_data.csv") 

### 2024-12-04 scripts/03-clean_data.R

@@ -1,33 +1,0 @@
-#### Preamble ####
-# Purpose: Clean the raw data recorded by WHO, removing unnecessary columns, and save it as csv and parquet.
-# Author: Manjun Zhu
-# Date: 26 November 2024
-# Contact: karmen.zhu@mail.utoronto.ca
-# License: MIT
-# Pre-requisites: finish raw-data
-
-#### Workspace setup ####
-library(tidyverse)
-library(arrow)
-
-#### Clean data ####
-data <- read_csv("data/01-raw_data/raw_data.csv")
-
-analysis_data <- data %>%
-  # Select the desired columns
-  select(Country, Year, Status, Life.expectancy, Diphtheria, 
-         Total.expenditure, Schooling, pct_doctor, pct_nursing) %>%
-  # Filter rows where Year is between 2009 and 2015, and drop the unrealistic value
-  filter(Year >= 2004 & Year <= 2015) %>%
-  filter(pct_doctor >= 0 & pct_doctor <= 1500) %>%
-  filter(pct_nursing >= 0 & pct_nursing <= 1500) %>%
-  # Remove rows with any missing values
-  drop_na()
-
-#### Rename column
-colnames(analysis_data)[colnames(analysis_data) == "Life.expectancy"] <- "LifeExpectancy"
-colnames(analysis_data)[colnames(analysis_data) == 
-                 "Total.expenditure"] <- "TotalExpenditure"
-
-#### Save data ####
-write_parquet(analysis_data, "data/02-analysis_data/analysis_data.parquet")

### 2024-12-04 scripts/04-test_analysis_data.R

@@ -1,103 +1,0 @@
-#### Preamble ####
-# Purpose: Performs a series of validation tests on cleaned Life Expectancy Data
-# Author: Manjun Zhu
-# Date: 26 November 2024
-# Contact: karmen.zhu@mail.utoronto.ca
-# License: MIT
-# Pre-requisites: The `tidyverse` , `arrow`, and `testthat` packages must be installed.
-# Any other information needed? Make sure you are in the `lifeexpectancydata` rproj
-
-
-#### Workspace setup ####
-library(tidyverse)
-library(testthat)
-library(arrow)
-
-analysis_data <- read_parquet(here::here("data/02-analysis_data/analysis_data.parquet"))
-
-
-#### Test data ####
-# Create list for results
-test_results <- list()
-
-# Test structure of data
-test_results$structure <- test_that("Data structure", {
-  expect_true(is.data.frame(analysis_data), 
-              info = "analysis_data is not a data frame.")
-  expect_equal(ncol(analysis_data), 10, info = "analysis_data does 
-               not have the correct number of columns.")
-  expect_equal(nrow(analysis_data), 1021, info = "analysis_data does 
-               not have the correct number of rows.")
-})
-
-# Test if column names match expected names
-test_results$column_names_test <- test_that("Column names", {
-  expected_colnames <- c("Country", "Year", "Status", "LifeExpectancy", "GDP", 
-                         "Diphtheria", "TotalExpenditure", "BMI", "Schooling", 
-                         "pct_doctor", "pct_nursing")
-  expect_equal(names(analysis_data), expected_colnames, 
-               info = "Column names do not match expected names.")
-})
-
-# Test if certain columns contain NA values
-test_results$na_values_test <- test_that("NA values", {
-  expect_false(any(is.na(analysis_data$LifeExpectancy)), 
-               info = "LifeExpectancy contains NA values.")
-  expect_false(any(is.na(analysis_data$Diphtheria)), 
-               info = "Diphtheria contains NA values.")
-  expect_false(any(is.na(analysis_data$pct_doctor)), 
-               info = "pct_doctor contains NA values.")
-  expect_false(any(is.na(analysis_data$pct_nursing)), 
-               info = "pct_nursing contains NA values.")
-  expect_false(any(is.na(analysis_data$Schooling)), 
-               info = "Schooling contains NA values.")
-  expect_false(any(is.na(analysis_data$TotalExpenditure)), 
[…]

### 2024-12-04 scripts/05-exploratory_data_analysis.R

@@ -1,43 +1,0 @@
-#### Preamble ####
-# Purpose: This script performs exploratory data analysis (EDA) on the cleaned life expectancy data
-# Author: Manjun Zhu
-# Date: 26 November 2024
-# Contact: karmen.zhu@mail.utoronto.ca
-# License: MIT
-# Pre-requisites: The 'tidyverse', 'here', and 'lubridate' packages must be installed
-
-
-# load libraries
-library(tidyverse)
-library(here)
-library(lubridate)
-
-# read cleaned data
-data <- read_parquet(here::here("data/02-analysis_data/analysis_data.parquet"))
-
-# Summary of numerical variables
-summary_stats <- data %>%
-  summarise(
-    LifeExpectancy = mean(LifeExpectancy, na.rm = TRUE),
-    Diphtheria = mean(Diphtheria, na.rm = TRUE),
-    TotalExpenditure = mean(TotalExpenditure, na.rm = TRUE),
-    Status = mean(as.numeric(Status), na.rm = TRUE),
-    Schooling = mean(Schooling, na.rm = TRUE),
-    pct_doctor = mean(pct_doctor, na.rm = TRUE),
-    pct_nursing = mean(pct_nursing, na.rm = TRUE),
-  )
-
-print(summary_stats)
-
-
-# Status breakdown
-status_distribution <- data %>%
-  count(Status) %>%
-  mutate(percent = n / sum(n) * 100)
-
-print(status_distribution)
-
-# number of NA values in each column
-missing_data <- data |>
-  summarise(across(everything(), ~ sum(is.na(.))))
-missing_data

### 2024-12-04 scripts/06-model_data.R

@@ -1,43 +1,0 @@
-#### Preamble ####
-# Purpose: This script performs out-of-sample testing and calculates the RMSE for the linear model trained in the previous script.
-# Author: Manjun Zhu
-# Date: 26 November 2024
-# Contact: karmen.zhu@mail.utoronto.ca
-# License: MIT
-# Pre-requisites: The `tidyverse`, `caret`, and 'here' packages must be installed
-
-
-#### Workspace setup ####
-library(tidyverse)
-library(here)
-
-#### Read data ####
-# convert some columns to the appropriate data types
-data <- read_parquet(here::here("data/02-analysis_data/analysis_data.parquet"))
-data <- data %>%
-  mutate(
-    Country = as.factor(Country),
-    Year = as.numeric(Year),
-    LifeExpectancy = as.numeric(LifeExpectancy),
-    Status = as.factor(Status),
-    Diphtheria = as.numeric(Diphtheria),
-    TotalExpenditure = as.numeric(TotalExpenditure),
-    Schooling = as.numeric(Schooling),
-    pct_doctor = as.numeric(pct_doctor),
-    pct_nursing = as.numeric(pct_nursing)
-  )
-
-### Model data ####
-# linear model for Developed Countries' Life Expectancy based on socioeconomic factors
-lm_developed <- lm(LifeExpectancy ~  Diphtheria + Schooling + TotalExpenditure * pct_doctor * pct_nursing, 
-                   data = data %>% filter(Status == "Developed"))
-
-
-# linear model for Developing Countries' Life Expectancy based on socioeconomic factors
-lm_developing <- lm(LifeExpectancy ~ Diphtheria + Schooling + TotalExpenditure * pct_doctor * pct_nursing, 
-                    data = data %>% filter(Status == "Developing"))
-
-#### Save model ####
-saveRDS(lm_developed, here::here("models/lm_developed.rds"))
-saveRDS(lm_developing, here::here("models/lm_developing.rds"))
-

### 2024-12-04 scripts/07-replications.R

@@ -1,50 +1,0 @@
-#### Preamble ####
-# Purpose: This script performs out-of-sample testing and calculates the RMSE for the linear model trained in the previous script.
-# Author: Manjun Zhu
-# Date: 26 November 2024
-# Contact: karmen.zhu@mail.utoronto.ca
-# License: MIT
-# Pre-requisites: The `tidyverse`, `caret`, and 'here' packages must be installed
-
-#### Workspace setup ####
-library(tidyverse)
-library(caret)
-library(here)
-
-# Load the cleaned dataset (adjust the file path as needed)
-data <- read_parquet(here::here("data/02-analysis_data/analysis_data.parquet"))
-
-# Clean the data again to ensure consistency with the model
-data_cleaned <- data %>%
-  filter(!is.na(LifeExpectancy), !is.na(Diphtheria), !is.na(Schooling), 
-         !is.na(TotalExpenditure), !is.na(pct_doctor)) %>%
-  mutate(
-    LifeExpectancy = as.numeric(LifeExpectancy),
-    Diphtheria = as.numeric(Diphtheria),
-    Schooling = as.integer(Schooling),
-    TotalExpenditure = as.integer(TotalExpenditure),
-    pct_doctor = as.integer(pct_doctor)
-  )
-
-# Split the data into training (80%) and testing (20%) sets
-set.seed(955) # For reproducibility
-Index <- createDataPartition(data_cleaned$LifeExpectancy,
-                                  p = 0.8,
-                                  list = FALSE,
-                                  times = 1
-)
-
-train_data <- data_cleaned[Index, ]
-test_data <- data_cleaned[-Index, ]
-
-# Load the saved model
-finisher_model <- readRDS(here::here("models/lm_developed.rds"))
-
-# Perform out-of-sample prediction on the test set
-predicted <- predict(finisher_model, newdata = test_data)
-
-# Calculate the RMSE (Root Mean Squared Error)
-rmse <- sqrt(mean((predicted - test_data$LifeExpectancy)^2))
-
-# Print RMSE value
-cat("Root Mean Squared Error (RMSE) on the test data: ", round(rmse, 2), "\n")

### 2024-12-04 paper/paper.qmd
Add files via upload

@@
 title: "What to Expect with Life Expectancy?"
-subtitle: "Examining How Status, Healthcare Expenditure, Education, and Medical Infrastructure Influence Life Expectancy Across Developed and Developing Regions"
+subtitle: "Examining How Healthcare Expenditure, Education, and Medical Infrastructure Influence Life Expectancy Across Developed and Developing Regions"
 author: 
@@
 However, this growth has not been evenly distributed across the globe, with wealthier nations benefiting disproportionately.
-This research examines the impact of a nation’s economic standing and health conditions on life expectancy, exploring nuanced variables beyond those traditionally studied.
-By developing a linear model, the analysis identifies key relationships between economic and healthcare indicators and their influence on population longevity.
-The findings aim to inform health policy and economic planning, providing actionable insights to address disparities and promote equitable growth in life expectancy worldwide."
+This research examines the impact of a nation’s educational standing and health conditions on life expectancy, exploring potential variables beyond those traditionally studied.
+By developing a linear model, the analysis identifies key relationships between education and healthcare indicators and their influence on population longevity.
+The findings aim to inform health policy and education planning, providing actionable insights to address disparities and promote equitable growth in life expectancy worldwide."
 format: pdf
@@
 library(here)
+library(arrow)
 library(ggplot2)
@@
 library(ggridges)
+library(broom)
 
 # load data
-data <- read.csv(here::here("data/02-analysis_data/analysis_data.csv"))
+data <- read_parquet(here::here("data/02-analysis_data/analysis_data.parquet"))
 
@@
 However, this progress has not been distributed equitably across the globe.
-While nations with robust economies and advanced healthcare systems have achieved significant gains, others, particularly in underdeveloped regions, continue to face barriers such as inadequate infrastructure, persistent diseases, and economic instability.
+While nations with robust economies and developed healthcare systems have achieved significant gains, others, particularly in underdeveloped regions, continue to face barriers such as inadequate infrastructure, persistent diseases, and economic instability.
 Despite existing research on life expectancy determinants, gaps remain in understanding the nuanced relationships between economic standing, healthcare access, and other critical variables.
 
-This study aims to examine the factors influencing life expectancy across different nations by developing a linear model that incorporates both healthcare and economic variables.
+This study aims to examine the factors influencing life expectancy across different nations by developing a linear regression model that incorporates both healthcare and education variables.
 The response variable in the analysis is life expectancy, while the predictors include national status (Developed vs.
@@
 Another study extended this understanding by exploring the role of the Human Development Index (HDI), emphasizing the influence of public health initiatives, economic conditions, and education on life expectancy in low and medium-HDI nations [@Girum2018].
-Besides that, advanced econometric techniques, including panel cointegration analysis, have demonstrated a positive relationship between health care expenditure per capita and life expectancy, providing a robust foundation for integrating economic variables into this study [@He2020].
+Besides that, developed econometric techniques, including panel cointegration analysis, have demonstrated a positive relationship between health care expenditure per capita and life expectancy, providing a robust foundation for integrating economic variables into this study [@He2020].
 Additionally, a finding [@Hanushek2016] underscored the importance of combining both educational investments and government support in improving life expectancy, particularly in developing regions.
@@
 
-The analysis revealed that life expectancy in developed countries is generally higher and more stable, attributed to stronger healthcare systems, consistent access to medical services, and robust economic stability.
+The analysis reveals that life expectancy in developed countries is generally higher and more stable, attributed to stronger healthcare systems, consistent access to medical services, and robust economic stability.
 These factors are typically supported by well-established public health policies.
@@
 This study utilizes two primary datasets related to global health and life expectancy.
-The first dataset, World Health Statistics 2020 Complete, sourced from Kaggle [@world_health_statistics], provides comprehensive global health data, including variables such as GDP, life expectancy, mortality rates, and access to healthcare.
-The second dataset, Life Expectancy (WHO), also from Kaggle [@life_expectancy_who], focuses specifically on life expectancy trends and factors, offering a detailed breakdown by country, year, and socioeconomic variables.
+The first dataset, Life Expectancy and Healthy Life Expectancy [@world_health_statistics], provides comprehensive global health data, including variables such as GDP, life expectancy, mortality rates, and access to healthcare.
+The second dataset, Life Expectancy (WHO) [@life_expectancy_who], focuses specifically on life expectancy trends and factors, offering a detailed breakdown by country, year, and socioeconomic variables.
 
@@
 
-We performed data cleaning and analysis using R [@citeR], incorporating a range of packages for efficient data manipulation, visualization, and statistical modeling, leveraging packages such as **`tidyverse`** for data manipulation and visualization [@tidyverse], **`dplyr`** for streamlined data wrangling [@dplyr], **`ggplot2`** for creating high-quality plots [@ggplot2], **`ggridges`** for displaying the distribution of variables across different categories with ridge plots [@ggridges], **`here`** for managing file paths [@here], **`lubridate`** for handling date-related variables [@lubridate], **`kableExtra`** for generating polished tables [@kableExtra], **`modelsummary`** for summarizing model results [@modelsummary].
+We performed data cleaning and analysis using R [@citeR], incorporating a range of packages for efficient data manipulation, visualization, and statistical modeling, leveraging packages such as **`tidyverse`** for data manipulation and visualization [@tidyverse], **`dplyr`** for streamlined data wrangling [@dplyr], **`ggplot2`** for creating high-quality plots [@ggplot2], **`ggridges`** for displaying the distribution of variables across different categories with ridge plots [@ggridges], **`ggbeeswarm`** for visualizing the distribution of data points with swarming plots [@ggbeeswarm], **`here`** for managing file paths [@here], **`lubridate`** for handling date-related variables [@lubridate], **`arrow`** for data storage and processing [@arrow], **`kableExtra`** for generating polished tables [@kableExtra], **`modelsummary`** for summarizing model results [@modelsummary],  **`broom`** for tidying up model outputs into easily readable data frames [@broom], and **`knitr`** for dynamic report generation and the integration of code with results in reproducible workflows [@knitr].
 
@@
 
-The measurement process refers to how real-world phenomena—such as healthcare access, economic indicators, or demographic factors—are translated into numerical entries in the dataset.
-Each entry corresponds to a specific country and year, capturing variables related to life expectancy and associated determinants.
+The measurement process explains how real-world phenomena—such as healthcare access, economic indicators, or demographic factors—are translated into structured data entries.
+Each row in the dataset represents a specific country in a given year, capturing various variables related to life expectancy and its determinants.
+These variables originate from data collection efforts by global organizations like the World Health Organization (WHO), which rely on surveys, national health systems, and standardized reporting frameworks to compile accurate and comparable statistics across countries.
 Below is a description of key variables in the dataset:
@@
 
-**Health Expenditure (`TotalExpenditure`)**: This represents the general government expenditure on health as a percentage of total government expenditure.
+**Percentage of Health Expenditure (`HealthExpenditure`)**: This represents the general government expenditure on health as a percentage of total government expenditure.
 It is reported by national governments or international organizations and recorded as a percentage in the dataset.
@@
 #| label: fig-response
-#| fig-cap: "Distribution of life expectancy across different Status groups (Developed vs.
-Developing), highlighting the distinct patterns in their density.
-The mode of the density for life expectancy in developed countries tends to be around 81 years with a narrower variance, whereas in developing countries, the mode is approximately 74 years with a wider variance, reflecting disparities in health outcomes."
+#| fig-cap: "The distribution highlights the frequency of life expectancy values across the dataset.
+The distribution is slightly right-skewed, with a mean around 73 years, reflecting the overall health trends captured in the data."
 #| echo: false
@@
   geom_histogram(aes(y = ..density..), 
-                 bins = 70, fill = "brown", color = "white", alpha = 0.7) +
-  geom_density(color = "royalblue", fill = "lightblue", alpha = 0.6) +
+                 bins = 70, fill = "#433769", color = "white", alpha = 0.7) +
+  geom_density(color = "#8980a8", fill = "#d0c7eb", alpha = 0.6) +
   labs(
-    title = "Distribution of Life Expectancy by Status",
-    x = "Life Expectancy", y = "Density"
+    title = "Distribution of Life Expectancy",
+    x = "Life Expectancy", y = "Frequency"
   ) +
-  theme_minimal() +
-  facet_wrap(~Status)
+  theme_minimal()
 
@@
 **Diphtheria Immunization Coverage (`Diphtheria`)**: This variable represents the percentage of one-year-olds who have received the Diphtheria, Tetanus, and Pertussis (DTP3) immunization.
-Immunization coverage(@fig-diphtheria) is an important health indicator, as it reflects the effectiveness of a country's healthcare system in preventing infectious diseases, which in turn affects life expectancy.
+Immunization coverage[@fig-diphtheria] is an important health indicator, as it reflects the effectiveness of a country's healthcare system in preventing infectious diseases, which in turn affects life expectancy.
 
@@
 4.
-**Medical Doctors (`pct_doctor`)**: This variable indicates the number of medical doctors per 10,000 people (@fig-doctor).
+**Medical Doctors (`pct_doctor`)**: This variable indicates the number of medical doctors per 10,000 people[@fig-doctor].
 Access to healthcare professionals is a key determinant of health outcomes, and a higher number of doctors generally corresponds with better healthcare services and longer life expectancy.
@@
 5.
-**Nursing and Midwifery Personnel (`pct_nursing`)**: Similar to the number of doctors, this variable represents the number of nursing and midwifery personnel per 10,000 population (@fig-nursing).
-Nurses play a crucial role in delivering healthcare, and higher availability of nursing staff can contribute to improved health outcomes and increased life expectancy.
+**Nursing and Midwifery Personnel (`pct_nursing`)**: Similar to the number of doctors, this variable represents the number of nursing and midwifery personnel per 10,000 population[@fig-nursing].
+Nurses play a significant role in delivering healthcare, and higher availability of nursing staff can contribute to improved health outcomes and increased life expectancy.
 
 7.
-**Health Expenditure (`TotalExpenditure`)**: This variable reflects the general government expenditure on health as a percentage of total government expenditure, as shown in @fig-totalExpenditure.
+**Percentage of Health Expenditure (`HealthExpenditure`)**: This variable reflects the general government expenditure on health as a percentage of total government expenditure, as shown in @fig-totalExpenditure.
 Higher health spending is often associated with better healthcare infrastructure, which can lead to longer life expectancy due to better disease prevention and treatment options.
@@
 #| label: fig-status
-#| fig-cap: "Histogram of Status, showing the counts of developed and developing countries in the dataset. There are 305 data points from developed countries and 788 from developing countries."
+#| fig-cap: "The distribution of Status shows the counts of developed and developing countries in the dataset. There are 305 data points from developed countries and 788 from developing countries."
 #| echo: false
@@
   geom_bar(stat = "count") +
-  geom_text(stat = "count", aes(label = ..count..), vjust = 1.5, size = 5) +  # Move text to the bottom
   labs(x = "Country Status", 
-       y = "Count", 
+       y = "Frequency", 
        title = "Distribution of Countries by Status (Developed vs Developing)") +
-  scale_fill_manual(values = c("royalblue", "orange")) +
+  scale_fill_manual(values = c("#b3685d", "#a85b6a")) +
   theme_minimal() +
@@
 #| label: fig-diphtheria
-#| fig-cap: "The plot shows similar immunization coverage conditions across developed and developing countries, with a noticeable increase in variance for developing countries due to some countries exhibiting low diphtheria immunization rates."
+#| fig-cap: "The distribution of immunization coverage for diphtheria reveals that immunization coverage is generally high, with most countries achieving high vaccination rates.
+However, there is some variance, reflecting disparities in immunization efforts across different regions."
 #| echo: false
@@
 
-# Calculate variance for each Status group
-variance_data <- data %>%
-  group_by(Status) %>%
-  summarise(variance = var(Diphtheria, na.rm = TRUE))
+data_clean <- data %>%
+  filter(!is.na(Diphtheria) & Diphtheria > 0)
 
-# Plot
-ggplot(data, aes(x = Diphtheria, y = Status, fill = Status)) +
-  geom_density_ridges(alpha = 0.7) +
-  # Add text for Variance on the left side
-  geom_text(data = variance_data, 
-            aes(x = 5, y = Status, label = paste("Variance:", round(variance, 1))),
-            size = 3, 
-            hjust = 0, 
-            color = "brown") +
-  labs(x = "Diphtheria Immunization Coverage (%)", 
-       y = "Country Status", 
-       title = "Distribution of Diphtheria Immunization by Country Status") +
-  scale_fill_manual(values = c("royalblue", "darkorange")) +
-  scale_x_continuous(limits = c(0, 100)) +  # Ensure x-axis is between 0 and 100
-  theme_minimal() +
-  theme(legend.position = "none")
+ggplot(data_clean, aes(x = Diphtheria)) +
+  geom_histogram(aes(y = ..density..), 
+                 bins = 70, fill = "#4d1f0a", color = "white", alpha = 0.7) +
+  labs(
+    title = "Distribution of Immunization Coverage of Diphtheria",
+    x = "Immunization Coverage of Diphtheria", y = "Frequency"
+  ) +
+  theme_minimal()
+
 ```
@@
 #| label: fig-schooling
-#| fig-cap: "Violin plot with boxplot for Schooling by Status, highlighting a narrow distribution for developed regions with a higher median of 16 years of schooling.
-In contrast, developing countries show a wider variance and a relatively lower median of around 12.5 years, indicating significant disparities in educational access and attainment, with developed countries having more consistent and higher levels of education compared to the varied and generally lower levels in developing countries."
+#| fig-cap: "The distribution of Schooling Years is shown as a representation of the Education Factor, the distribution of schooling years across the dataset approaches a normal distribution"
 #| echo: false
@@
 
-ggplot(data, aes(
-  x = Status,
-  y = Schooling,
-  fill = Status
-)) +
-  geom_violin(alpha = 0.5, position = position_dodge(width = 1)) + 
-  geom_boxplot(width = 0.1, color = "white", outlier.shape = NA, position = position_dodge(width = 1)) +
+data_clean <- data %>%
+  filter(!is.na(Schooling) & Schooling > 0)
+
+ggplot(data_clean, aes(x = Schooling)) +
+  geom_histogram(aes(y = ..density..), 
+                 bins = 15, fill = "#629596", color = "white", alpha = 0.7) +
   labs(
-    x = "Status",
-    y = "Schooling Years",
-    title = "Distribution of Schooling Years by Status"
+    title = "Distribution of Schooling Years",
+    x = "Schooling Years", y = "Frequency"
   ) +
-  scale_fill_manual(values = c("Developed" = "brown", "Developing" = "darkgreen")) + 
-  # Separate panels for each status
-  facet_wrap(~Status, scales = "free_x") +
-  theme_minimal() +
-  theme(axis.text.x = element_text(angle = 45, hjust = 1))
+  theme_minimal()
 
@@
 #| label: fig-totalExpenditure
-#| fig-cap: "Boxplot for government expenditure on health as a percentage of total expenditure, showing relatively high percentages for developed countries compared to developing countries, highlighting that developed countries' governments typically allocate more resources to health, though there are exceptions."
+#| fig-cap: "The distribution of government expenditure on health as a percentage of total expenditure reflects the level of medical support provided by governments."
 #| echo: false
@@
 
-ggplot(data, aes(x = Status, y = TotalExpenditure, fill = Status)) +
-  # Adding boxplots on top of violin plots for more detail on median and quartiles
-  geom_boxplot(width = 0.5, alpha = 0.7, position = position_dodge(0.9)) +
+data_clean <- data %>%
+  filter(!is.na(TotalExpenditure) & TotalExpenditure > 0)
 
+ggplot(data_clean, aes(x = TotalExpenditure)) +
+  geom_histogram(aes(y = ..density..), 
+                 bins = 20, fill = "#5765ad", color = "white", alpha = 0.7) +
   labs(
-    title = "Total Government Health Expenditure by Status",
-    x = "Country Status",
-    y = "Total Government Expenditure on Health (%)"
+    title = "Distribution of Government Expenditure on Health",
+    x = "Percentage of Health Expenditure as a Percentage of Total Expenditure", y = "Frequency"
   ) +
-  theme_minimal() +
-  scale_fill_manual(values = c("skyblue", "orange")) +  # Customize colors for Developed and Developing
-  theme(legend.position = "none")  # Hide the legend for clarity
+  theme_minimal()
 
@@
 #| label: fig-doctor
-#| fig-cap: "Density plot for doctors per 10,000 population by Status, showing that the density of medical doctors in developed regions is relatively higher compared to developing regions.
-In developing countries, the density is concentrated at lower levels, suggesting that the healthcare system is generally stronger and more accessible in developed countries."
+#| fig-cap: "The distribution of doctors per 10,000 population, providing an overview of the distribution of medical doctors globally, reflecting general medical resource availability."
 #| echo: false
@@
 
-ggplot(data, aes(x = pct_doctor, y = Status, fill = Status)) +
-  geom_density_ridges(alpha = 0.7) +
-  labs(x = "Density of Medical Doctor", y = "Country Status", title = "Medical Doctor Condition") +
-  scale_fill_manual(values = c("royalblue", "darkorange")) +
-  theme_minimal() +
-  theme(legend.position = "none")
+ggplot(data, aes(x = pct_doctor)) +
+  geom_histogram(aes(y = ..density..), 
+                 bins = 15, fill = "#8b6fb0", color = "white", alpha = 0.7) +
+  labs(
+    title = "Distribution of the number of doctors per 10,000",
+    x = "Number of medical doctors per 10,000 population", 
+    y = "Frequency"
+  ) +
+  theme_minimal()
 
@@
 #| label: fig-nursing
-#| fig-cap: "Number of Nurses by Status with Highlighted Mean and Dotplot.
-The mean for developed countries is around 80, while the mean for developing countries is around 35, with many data points showing low nurse density in developing countries.
-Both status groups have medians lower than their means, suggesting a right-skewed distribution with a concentration of countries with fewer nurses."
+#| fig-cap: "The distribution of nurses per 10,000 population, offering a broad view of nurse density worldwide as an indicator of healthcare support."
 #| echo: false
@@
 
-ggplot(data, aes(x = Status, y = pct_nursing, fill = Status)) +
-  geom_boxplot(alpha = 0.8, outlier.shape = NA) +  # Boxplot for the distribution
-  geom_dotplot(binaxis = "y", stackdir = "center", dotsize = 0.5, alpha = 0.4, position = "dodge") +  # Smaller dots with dodge position
-  stat_summary(fun = "mean", geom = "point", shape = 18, size = 4, color = "black", fill = "yellow") +  # Highlight the mean
+ggplot(data, aes(x = pct_nursing)) +
+  geom_histogram(aes(y = ..density..), 
+                 bins = 15, fill = "#3f594a", color = "white", alpha = 0.8) +
   labs(
-    title = "Number of Nurses by Status with Highlighted Mean and Dotplot",
-    x = "Status",
-    y = "Number of Nurses per 10000"
+    title = "Distribution of the number of nurses per 10000",
+    x = "the number of nurses and  per 10000 population", y = "Frequency"
   ) +
-  theme_minimal() +
-  scale_fill_manual(values = c("brown", "darkgreen")) +  # Customize fill colors
-  theme(axis.text.x = element_text(angle = 45, hjust = 1))  # Tilt x-axis labels
+  theme_minimal()
 
@@
 Define $y_i$ as the Life Expectancy for country $i$.
-The factors $x_1$, $x_2$,...represent socioeconomic predictors such as Diphtheria, Schooling, TotalExpenditure, pct_doctor, pct_nursing, and interaction term.
+The factors $x_1$, $x_2$,...represent socioeconomic predictors such as Diphtheria Immunization Coverage, Schooling, Health Expenditure, Number of Doctors and Nurses per 10000, and interaction term.
 
@@
 
-Linear regression assumes that the residuals errors of the model are normally distributed.
-If this assumption is violated, it could lead to biased estimates and incorrect conclusions.
-This issue is especially important in cross-country data, where life expectancy may be influenced by diverse factors that do not follow a normal distribution.
+The linear regression model used to estimate life expectancy relies on several key assumptions that must be met for valid results.
+First, it assumes linearity, meaning there is a straight-line relationship between the predictors (e.g., healthcare condition, education levels) and the outcome variable (`life expectancy`).
+This assumption ensures that a change in a predictor corresponds to a consistent change in life expectancy.
+Second, the model assumes that the errors are uncorrelated, meaning the residuals do not exhibit any patterns when plotted.
+This assumption is significant for making accurate inferences, such as significance tests and confidence intervals.
+Third, the model assumes homoscedasticity, which means that the variance of the residuals should remain constant across all values of the predictors.
+This assumption helps avoid bias in estimating the relationship between the predictors and the outcome variable.
+Fourth, the residuals should follow a normal distribution.
+This assumption is necessary for performing accurate hypothesis tests and for generating valid confidence intervals, which provide insights into the uncertainty of the model's predictions.
+Lastly, it is important that the model does not appear multicollinearity, where predictors are highly correlated with each other.
+These assumptions provide a foundation for the model, violations can lead to misleading results with inaccurate coefficients.
 
-Predictors such as the number of doctors per 10,000 and the number of nurses per 10,000 might be correlated with each other.
-High correlation between predictors can cause multicollinearity, making it difficult to assess the individual contribution of each predictor.
+However, several limitations apply to this model.
+Despite the assumption of linearity, real-world relationships between life expectancy and its predictors may be more complex or non-linear, and such complexity would not be fully captured by this model.
+For instance, the effect of economic development on life expectancy may vary significantly at different levels of income.
+Moreover, while the model assumes uncorrelated errors, there may be factors that influence both the predictors and life expectancy, leading to correlated residuals.
+For example, social determinants like income inequality or access to healthcare might impact both education and life expectancy, potentially violating the assumption of independence.
+The assumption of constant variance also has limitations, as certain groups (e.g., countries with lower health expenditure) may have more variability in life expectancy than others, potentially leading to not constant variance.
+Additionally, the assumption of normally distributed errors may not hold in all cases, especially when extreme outliers (such as countries with unusual life expectancy rates) are present in the data.
+
+Other external factors not accounted for in this model may also influence life expectancy.
+Environmental factors like climate change, sudden public health crises (e.g., pandemics), and political instability could have significant effects on life expectancy, yet these factors may not be included as predictors in the model.
+These limitations should be considered when interpreting the model’s results, particularly when applying the findings to specific populations or making broad predictions.
 
@@
 In @tbl-keyestimates, the Developed model is more efficient and better fitted primarily due to its higher R² (64.5% vs 51.5%), indicating that it explains a greater proportion of variance in life expectancy.
-Additionally, its RMSE value is lower (3.78 vs 5.06), showing more precise predictions.
+Additionally, its RMSE value is lower (3.76 vs 4.94), showing more precise predictions.
 The Developing model has a larger sample size and variability, which might have diluted the effect of individual predictors.
@@
 
-### Model for Developed Countries
-
-For the Developed model, the predictors considered included Diphtheria Immunization, Schooling, Total Expenditure, and Number of Doctors per 10000.
-The coefficients and standard errors for these predictors were as follows:
-
-**Diphtheria Immunization**: Coefficient = 0.022, Standard Error = 0.025
-
-**Schooling**: Coefficient = 0.801, Standard Error = 0.151
-
-**Health Expenditure**: Coefficient = 0.056, Standard Error = 0.084
-
-**Number of Medical Doctors per 10000**: Coefficient = 0.018, Standard Error = 0.034
-
-These coefficients indicate that `Schooling` has the highest contribution to explaining life expectancy, followed by `Diphtheria Immunization`.
-The relatively small contribution of `#Doctors per 10000` suggests that doctor practices may have a less direct impact on life expectancy in developed countries, but this could be nuanced by other health factors not captured in the model.
-
-### Model for Developing Countries
-
-For the Developing model, the same predictors were considered, but their relationships with life expectancy differed. The coefficients and standard errors for each predictor were as follows:
-
-**Diphtheria Immunization**: Coefficient = 0.036, Standard Error = 0.010
-
-**Schooling**: Coefficient = 1.727, Standard Error = 0.113
-
-**Number of Medical Doctors per 10000**: Coefficient = 0.090, Standard Error = 0.021
-
-**Number of Nurses per 10000**: Coefficient = -0.003, Standard Error = 0.003
-
-In the Developing model, `Schooling` again has the most significant impact, and `#Doctors per 10000` is also a substantial predictor.
-This highlights the importance of healthcare workforce availability in improving life expectancy in developing countries, unlike the developed context where Schooling was more prominent.
-The negative relationship of `Total Expenditure` suggests that in developing countries, spending in certain healthcare sectors could be misallocated or inefficient.
-```{r}
-#| label: tbl-keyestimates
-#| tbl-cap: "Summary of key model estimates for Developed and Developing counties, including coefficients for predictors like Diphtheria, Schooling, and TotalExpenditure, pct_doctor, pct_nursing, with standard errors for each estimate.
-Model performance statistics, such as sample size, R², and adjusted R², are also displayed."
-#| echo: false
-#| eval: true
-#| warning: false
-#| message: false
-
-# load data and model
-lm_developed <- readRDS(here::here("models/lm_developed.rds"))
-lm_developing <- readRDS(here::here("models/lm_developing.rds"))
-
-model_list <- list("Model for Developed" = lm_developed, "Model for Developing" = lm_developing)
-modelsummary(model_list, 
-             output = "kableExtra", 
-             statistic = "std.error", 
-             coef_map = c("Diphtheria" = "Diphtheria Immunization",
-                          "Schooling" = "Schooling",
-                          "TotalExpenditure" = "Health Expenditure",
-                          "pct_doctor" = "Number of Medical Doctors per 10000",
-                          "pct_nursing" = "Number of Nurses per 10000")) %>%
-  kable_styling(latex_options = c("striped", "scale_down"))
-```
-
-\newpage
+Extended versions of these tables can be found in @tbl-additionaltable
 
@@
 #| label: tbl-modelresults
-#| tbl-cap: "The prediction table indicates that the average life expectancy in developed countries is 78.53 years, while in developing countries, it is slightly lower at 71.68 years.
+#| tbl-cap: "The prediction table indicates that the average life expectancy in developed countries is 78.52 years, while in developing countries, it is slightly lower at 71.67 years.
 This difference of approximately 5.1 years in life expectancy suggests a significant health and infrastructure disparity between developed and developing countries.
-These results align with broader global trends where developed countries typically benefit from better healthcare systems, higher standards of living, and advanced medical technology, contributing to longer life expectancies.
+These results align with broader global trends where developed countries typically benefit from better healthcare systems, higher standards of living, and developed medical technology, contributing to longer life expectancies.
 "
@@
 # load data and model
-data <- read.csv(here::here("data/02-analysis_data/analysis_data.csv"))
 lm_developed <- readRDS(here::here("models/lm_developed.rds"))
@@
   )
+
+```
+
+```{r}
+#| label: tbl-keyestimates
+#| tbl-cap: "Summary of key model estimates for Developed and Developing counties, including coefficients for predictors like Diphtheria, Schooling, and TotalExpenditure, pct_doctor, pct_nursing, with standard errors for each estimate.
+Model performance statistics, such as sample size, R², and adjusted R², are also displayed."
+#| echo: false
+#| eval: true
+#| warning: false
+#| message: false
+
+# load data and model
+lm_developed <- readRDS(here::here("models/lm_developed.rds"))
+lm_developing <- readRDS(here::here("models/lm_developing.rds"))
+
+model_list <- list("Model for Developed" = lm_developed, "Model for Developing" = lm_developing)
+modelsummary(model_list, 
+             output = "kableExtra", 
+             statistic = "std.error", 
+             coef_map = c("Diphtheria" = "Diphtheria Immunization",
+                          "Schooling" = "Schooling",
+                          "TotalExpenditure" = "Health Expenditure",
+                          "pct_doctor" = "Number of Medical Doctors per 10000",
+                          "pct_nursing" = "Number of Nurses per 10000")) %>%
+  kable_styling(latex_options = c("striped", "scale_down"))
 ```
@@
 
-data <- read.csv(here::here("data/02-analysis_data/analysis_data.csv"))
 lm_developed <- readRDS(here::here("models/lm_developed.rds"))
@@
     PredictedDeveloped = predict(lm_developed, newdata = data),
-    PredictedDeveloping = predict(lm_developing, newdata = data)
+    PredictedDeveloping = predict(lm_developing, newdata = data),
   )
@@
 ```
+
+\newpage
+
+The predicted life expectancy distributions across various predictors reveal important trends.
+Longer schooling years[@fig-schoolgroups] strongly correlate with higher predicted life expectancy, emphasizing the positive impact of education.
+High diphtheria immunization coverage[@fig-diphtheria-groups] is also associated with increased life expectancy, though this relationship is less consistent for medium and low coverage levels, indicating variability in its predictive contribution.
+Similarly, higher government health expenditure[@fig-totalexpenditure-groups] aligns with greater life expectancy, while lower expenditure shows limited variability in predicted outcomes.
+Lastly, the distribution of medical professionals[@fig-doctornursing-groups] highlights that developed medical conditions generally contribute to higher life expectancy, though some variability in their influence remains.
+
+```{r}
+#| echo: false
+#| eval: true
+
+# Load the model
+lm_overall <- readRDS(here::here("models/lm_overall.rds"))
+
+# Generate predictions
+prediction_data <- data %>%
+  filter(!is.na(Diphtheria) & !is.na(TotalExpenditure) & !is.na(Schooling) & !is.na(pct_doctor) & !is.na(pct_nursing))
+
+prediction_data$predictedLifeExpectancy <- predict(lm_overall, newdata = prediction_data)
+
+```
+
+```{r}
+#| echo: false
+#| eval: true
+#| label: fig-schoolgroups
+#| fig-cap: "The distribution shows how different levels of education correlate with predicted life expectancy, highlighting group-wise distribution and variability."
+#| warning: false
+#| message: false
+
+# Remove rows with 0 or NULL values in Schooling
+prediction_data <- prediction_data %>%
+  filter(!is.na(Schooling) & Schooling > 0)
+
+# Create Schooling year groups for segmentation
+prediction_data$schooling_years <- cut(
+  prediction_data$Schooling,
+  breaks = c(3.1, 6.1, 12.1, 15.1, 18.1, 21),
+  labels = c("3-6", "6-12", "12-15", "15-18", "18-21")
+)
+
+# Violin plot
+ggplot(prediction_data, aes(x = schooling_years, y = predictedLifeExpectancy, fill = schooling_years)) +
+  geom_violin(alpha = 0.6, trim = FALSE) + # Violin plot for distribution
+  scale_fill_manual(
+    values = c("3-6" = "blue", "6-12" = "green", "12-15" = "orange", "15-18" = "purple", "18-21" = "red"),
+    name = "Schooling Year Groups"
+  ) +
+  labs(
+    title = "Predicted Life Expectancy by Schooling Year Groups",
+    x = "Schooling Year Groups",
+    y = "Predicted Life Expectancy (years)"
+  ) +
+  theme_minimal()
+
+```
+
+```{r}
+#| echo: false
+#| eval: true
+#| label: fig-diphtheria-groups
+#| fig-cap: "The distribution illustrates how varying immunization coverage levels contribute to differences in predicted life expectancy."
+#| warning: false
+#| message: false
+
+# Create Diphtheria coverage groups
+prediction_data$Diphtheria_groups <- cut(prediction_data$Diphtheria,
+  breaks = c(0, 30, 60, 100),
+  labels = c("0-30", "30-60", "60-100")
+)
+
+# Plot predicted life expectancy by Diphtheria groups using density plot
+ggplot(prediction_data, aes(x = predictedLifeExpectancy, fill = Diphtheria_groups)) +
+  geom_density(alpha = 0.7, adjust = 1.5) +  # Density plot to show distribution
+  scale_fill_manual(
+    values = c("0-30" = "royalblue", "30-60" = "lightgreen", "60-100" = "orange"),
+    name = "Diphtheria Coverage Groups"
+  ) + # Custom colors for each group
+  labs(
+    title = "Predicted Life Expectancy Across Diphtheria Coverage Groups",
+    x = "Predicted Life Expectancy (years)",
+    y = "Density"
+  ) +
+  theme_minimal() +
+  theme(legend.position = "right")  # Place legend on the right
+
+```
+
+```{r}
+#| echo: false
+#| eval: true
+#| label: fig-totalexpenditure-groups
+#| fig-cap: "The distribution reveals the distribution of predicted life expectancy across health expenditure levels, showing central tendencies and variability."
+#| warning: false
+#| message: false
+
+# Group TotalExpenditure
+prediction_data$TotalExpenditure_groups <- cut(
+  prediction_data$TotalExpenditure,
+  breaks = c(1.1, 4.26, 6.1, 7.98, 14.4),
+  labels = c("1.1-4.26", "4.26-6.1", "6.1-7.98", "7.98+")
+)
+
+# Boxplot for TotalExpenditure groups with empty fill and colored stroke
+ggplot(prediction_data, aes(x = TotalExpenditure_groups, y = predictedLifeExpectancy, color = TotalExpenditure_groups)) +
+  geom_boxplot(
+    fill = "white",  # Set the box fill to be empty (white)
+    alpha = 0.7     # Adjust transparency for the box outline
+  ) +
+    stat_summary(fun = "mean", geom = "point", shape = 16, size = 2, color = "brown") +  # Highlight the mean
+  scale_color_manual(
+    values = c("1.1-4.26" = "purple", "4.26-6.1" = "royalblue", "6.1-7.98" = "orange", "7.98+" = "lightcoral"),
+    name = "Health Expenditure Groups"
+  ) +
+  labs(
+    title = "Predicted Life Expectancy by Health Expenditure Groups",
+    x = "Health Expenditure Groups",
+    y = "Predicted Life Expectancy (years)"
+  ) +
+  theme_minimal() +
+  theme(legend.position = "right")  # Place legend on the right
+
+```
+
+```{r}
+#| echo: false
+#| eval: true
+#| label: fig-doctornursing-groups
+#| fig-cap: "The distribution demonstrates the relationship between medical professional availability (doctors and nurses) and predicted life expectancy, highlighting patterns across groups."
+#| warning: false
+#| message: false
+
+# Categorize predictedLifeExpectancy based on min, median, mean, and max values
+prediction_data$life_expectancy_group <- cut(prediction_data$predictedLifeExpectancy,
+  breaks = c(42.3, 60, 72.9, 80, 89),  # Define the breaks
+  labels = c("42.3-60", "60-72.9", "72.9-80", "80-89"),  # Labels for each group
+  include.lowest = TRUE
+)
+
+# Plot with custom colors for the life expectancy groups
+ggplot(prediction_data, aes(x = pct_doctor, y = pct_nursing, color = life_expectancy_group)) +
+  geom_point(alpha = 0.7, size = 2) +  # Scatter plot with points, color based on life expectancy groups
+  scale_color_manual(values = c("42.3-60" = "brown", "60-72.9" = "darkgreen", "72.9-80" = "orange", "80-89" = "purple")) +  # Custom colors for each life expectancy group
+  labs(
+    title = "Predicted Life Expectancy by Pct Doctor and Pct Nursing",
+    x = "the number of doctors per 10000",
+    y = "the number of nurses per 10000",
+    color = "Life Expectancy Groups"
+  ) +
+  theme_minimal() +
+  theme(legend.position = "right")  # Position the legend on the right side
+
+```
+
 \newpage
@@
 
-## First discussion point {#sec-first-point}
+## Factors Influencing Life Expectancy Across Nations {#sec-first-point}
 
-If my paper were 10 pages, then should be be at least 2.5 pages. The discussion is a chance to show off what you know and what you learnt from all this.
+The study aims to develop a model for predicting life expectancy based on healthcare, economic, and educational factors.
+By analyzing the relationship between these predictors and actual life expectancy, we gain insights into the factors that influence life expectancy outcomes and evaluate the model's predictive performance.
 
-## Second discussion point
+A comparison of predicted life expectancy distributions reveals several notable trends.
+Longer schooling years[@fig-schoolgroups] strongly correlate with higher life expectancy, highlighting the critical role of education.
+High diphtheria immunization coverage[@fig-diphtheria-groups] generally predicts greater life expectancy, though medium and low coverage levels exhibit variability, suggesting that immunization is not the sole or dominant factor.
+Similarly, higher government health expenditure[@fig-totalexpenditure-groups] aligns with improved outcomes, while lower spending shows minimal variation.
+Availability of doctors and nurses[@fig-doctornursing-groups] also indicates a positive relationship with life expectancy, though some variability suggests other contributing factors are at play.
 
-Please don't use these as sub-heading labels - change them to be what your point actually is.
+These findings align with prior research, demonstrating that healthcare and education are pivotal in shaping life expectancy.
+The model effectively captures these relationships, offering valuable insights for addressing disparities between developed and developing nations.
 
-## Third discussion point
+## Educational and Healthcare Factors as Significant Predictors
 
-## Weaknesses and next steps
+A key finding of this study is the distinct roles that educational and healthcare factors play in predicting life expectancy across developed and developing countries.
+In the model for developed nations, schooling is a significant positive predictor (0.645), though its impact is less pronounced compared to developing nations (1.694).
+This suggests that while education universally contributes to longer life expectancy, its influence is more significant in countries where educational access is still expanding.
 
-Weaknesses and next steps should also be included.
+Healthcare variables show contrasting trends between the two groups.
+Health expenditure negatively correlates with life expectancy in both contexts but is more substantial in developed countries (−1.740) compared to developing countries (−0.352).
+This reflects diminishing returns on investment in already developed healthcare systems in developed nations, where additional spending may not yield proportional gains in life expectancy.
+Similarly, the availability of medical doctors demonstrates a divergent impact, showing a negative association in developed nations (−0.446) but a positive one in developing nations (0.444).
+This suggests that increasing access to medical professionals is a more critical driver of life expectancy in developing regions, where healthcare systems are less established.
 
-\newpage
+These findings highlight the nuanced roles of education and healthcare in shaping life expectancy, emphasizing the need for tailored policy approaches to address disparities across nations.
 
-\appendix
+## Variance in Life Expectancy and the Influence of Social Factors
+
+The analysis underscores significant disparities in life expectancy variance between developed and developing countries, shaped by both systemic and social factors.
+In developed nations, life expectancy remains higher with less variance, a reflection of robust healthcare infrastructure, widespread access to medical services, and economic stability.
+These advantages are complemented by stronger social support systems, higher education levels, and improved public health awareness, which collectively enhance the consistency of health outcomes.
+
+In contrast, developing countries exhibit greater variability in life expectancy, driven by a complex interplay of economic growth, healthcare access, and social determinants such as education and community infrastructure.
+Regions with better access to schooling and healthcare services tend to show faster life expectancy improvements, while others lag behind due to inequities in resource distribution and social support.
+This dynamic highlights the dual role of economic and social factors in shaping life expectancy trends, underscoring the need for comprehensive policies that prioritize education, equitable healthcare, and community development to reduce disparities and foster sustainable progress.
+
+## Weaknesses and Next Steps
+
+One limitation of the current model is the exclusion of other potential factors that could influence life expectancy.
+While the model includes several key predictors like schooling, health expenditure, and the number of medical professionals, it does not account for other significant variables such as social determinants of health (e.g., income inequality, access to social services, or political stability), which could have a notable impact, particularly in developing countries.
+Environmental factors such as air quality, climate change, and water sanitation also play critical roles, especially in low-income regions, yet these were not included in the model.
+Additionally, there is the potential for bias in the dataset itself.
+If the dataset used for training the model is not representative of all countries, particularly those with unique socio-economic or healthcare challenges, it could limit the model’s generalizability.
+Bias could also arise if the data predominantly reflects developed nations, where healthcare systems and economic factors are more consistent, leading to skewed predictions for developing countries.
+
+Future research should explore several avenues to improve the model.
+First, expanding the dataset to include additional variables, such as policies related to social welfare, political stability, and government regulations, would provide a more comprehensive view of the factors influencing life expectancy.
+Incorporating data on policy changes or government reforms that impact healthcare and social services could also enrich the model's predictive power.
+Another area for future research could involve exploring the impact of environmental factors, such as pollution levels, access to clean water, and geographical conditions.
+These variables are especially relevant in developing countries and could reveal new insights into life expectancy trends.
+By addressing these limitations and exploring these additional variables, future research can refine models of life expectancy and improve predictions for countries at various stages of economic and healthcare development.
 
@@
 
-# Additional data details
+## Data cleaning {#sec-data-clean}
 
-# Model details {#sec-model-details}
+To prepare the dataset for analysis, I followed these steps:
 
-## Posterior predictive check
+1.
+Column Selection and Formatting: Removed irrelevant columns, such as administrative codes, and reformatted the "Year" column into a standardized four-digit format (YYYY) for consistency.
+Created a "Health_Care_Access" score by aggregating scaled values for "Medical Doctors per 10,000" and "Health Expenditure" to enable focused analysis.
 
-## Diagnostics
+2.
+Handling Missing Values and Grouping Categories: Addressed missing values in key variables like "Schooling" and "Health Expenditure" by applying group-wise median imputation based on development status (Developed or Developing).
+For categorical fields like "Status," missing entries were labeled as "Unknown." Countries with over 50% incomplete data were excluded, while less common regions were grouped into an "Other Regions" category to simplify comparisons.
+
+3.
+Adding New Columns: Created a "Health_Care_Access" score by aggregating scaled values for "Medical Doctors per 10,000" and "Nurses per 10,000" Added a binary "Developed_Status" column, encoding 1 for developed countries and 0 for developing countries, based on established World Bank classifications.
+
+4.
+Outlier Detection and Export: Verified and handled outliers in "Health Expenditure" and "Life Expectancy" using interquartile range (IQR) thresholds.
+The cleaned dataset was exported in both CSV and Parquet formats for efficient storage and compatibility with developed tools.
+
+## Idealized Methodology
+
+### **Population, Frame, and Sample**
+
+The population for this study encompasses all individuals worldwide whose life expectancy is influenced by various socioeconomic, healthcare, and environmental factors.
+This includes individuals living in developed and developing countries, spanning different ages, genders, and social groups.
+
+The frame consists of a comprehensive dataset of individuals with demographic, healthcare, and socioeconomic information relevant to life expectancy.
+This dataset could be sourced from global organizations such as the World Health Organization (WHO), World Bank, or national statistical agencies.
+The frame should ideally include recent data to capture current trends and factors affecting life expectancy.
+
+The sample would be a representative subset of individuals drawn from the frame, stratified by development status (Developed or Developing), age groups, and regions.
+This stratification ensures diversity and allows for comparisons across different demographics.
+The sample should include sufficient representation from high-variance populations, such as those in regions undergoing rapid economic or healthcare changes, to enhance the model's predictive accuracy.
+
+### **Sample Recruitment**
+
+-   **Recruitment via National and Global Surveys**: Individuals could be recruited through existing national and international health surveys, such as those conducted by the World Health Organization (WHO), World Bank, or national census bureaus.
+Participants would be invited to contribute additional data on healthcare access, education levels, and socioeconomic factors relevant to life expectancy.
+
+-   **Collaboration with Healthcare Providers**: Healthcare facilities, including hospitals and clinics, could assist in recruiting participants by inviting patients to contribute anonymized health data for research purposes.
+Recruitment could occur during routine health check-ups or through online health portals.
+
+-   **Community Outreach Programs**: Recruitment could target communities through local organizations, NGOs, and social programs.
+These efforts could focus on ensuring representation from underrepresented or high-variance populations, such as rural areas or rapidly developing regions.
+
+-   **Incentives**: Participants could be incentivized with benefits such as access to personalized health reports, free health screenings, or community-based health initiatives to improve overall participation rates.
+
+-   **Digital Platforms**: Online platforms like health-focused apps, educational websites, and social media groups could provide an avenue for researchers to engage potential participants and collect survey data.
+These platforms are especially useful for reaching younger or tech-savvy demographics.
+
+### **Sampling Approach**
+
+The study employs a stratified random sampling method, dividing the population into strata based on key factors that influence life expectancy, such as geographic region (e.g., developed vs.
+developing countries), age group (e.g., 0-20, 21-40, 41-60, 61+), income level (low, middle, high), and access to healthcare (e.g., urban, suburban, rural).
+This ensures that the sample represents diverse subgroups affecting life expectancy outcomes.
+For instance, suppose the study aims to collect data from 1,200 respondents.
+If stratification assigns 50% of the sample to developed countries and 50% to developing countries, with age distribution set at 20% for 0-20, 30% for 21-40, 30% for 41-60, and 20% for 61+, then 180 participants from developed countries would be aged 21-40.
+Further stratification could divide these respondents by income level, where 40% fall under the high-income bracket, equating to 72 participants.
+
+-   **Representativeness**: Stratified sampling ensures that the sample reflects the variability in life expectancy determinants, enabling more precise insights into how different factors (e.g., healthcare access, socioeconomic status) influence outcomes.
+
+-   **Precision**: By ensuring balanced representation from various subgroups, this approach minimizes biases and provides clearer insights into the differences between demographic or regional groups.
+
+-   **Complexity**: The need to define and categorize strata, particularly in diverse populations, can make stratified sampling more complex to implement.
+For example, defining healthcare accessibility uniformly across regions may be challenging.
+
+-   **Cost and Logistics**: Stratified sampling may involve higher costs and more administrative effort, particularly in reaching underrepresented or remote populations.
+Recruitment and data collection across strata could also take longer compared to simpler methods like random or convenience sampling.
+
+### **Non-response handling**
+
+Non-response occurs when individuals or groups do not provide their data, fail to complete surveys, or are otherwise unavailable for participation.
+Effectively managing non-response is significant to preserving the representativeness and reliability of the findings.
+
+-   **Follow-up Invitations**: If participants do not respond initially, reminders can be sent via email, phone calls, or community leaders.
+Personalized messages highlighting the importance of their input to public health and societal improvements may encourage participation.
+
+-   **Weighting for Non-Response**: Statistical weighting can be applied if certain demographics (e.g., rural residents, lower-income groups) have higher non-response rates.
+Adjusting weights ensures that these groups are proportionally represented in the analysis, mitigating potential biases.
+
+-   **Adjustment**: If systematic non-response is identified (e.g., fewer responses from regions with limited internet access), additional data collection efforts in these areas or adjustments during analysis using statistical techniques like imputation can be employed to fill gaps.
+
+-   **Engaging Local Organizations**: Partnering with local health agencies or NGOs in areas with traditionally low participation can foster trust and improve response rates.
+
+### **Standards for the Questionaires**
+
+-   **More Context**: Participants could be asked to provide detailed information on factors influencing life expectancy, such as diet, physical activity levels, access to healthcare, occupational hazards, and exposure to pollution.
+Including such variables could enhance the model's ability to capture key influences beyond demographic or geographic data.
+
+-   **Flexibility**: The questionnaire could utilize a mix of open-ended and scaled questions to gather both quantitative and qualitative insights.
+This design would allow participants to elaborate on unique life circumstances or local conditions that may not be captured in structured questions.
+
+-   **Self-Reporting Bias**: A significant limitation of questionnaires is the tendency for self-reporting bias.
+Participants might overestimate healthy behaviors, such as exercise or nutritious eating, or underreport harmful habits like smoking or alcohol consumption.
+This could distort the findings and reduce model accuracy.
+
+## Idealized Survey
+
+A sample survey can be found at: <https://forms.gle/Jzpw6MR5hXgnAfiDA>
+
+This survey is conducted as part of a research study to explore how various factors—such as healthcare condition, immunization coverage, and education—impact life expectancy, providing valuable insights into their relationship with overall health outcomes in your community.
+It includes questions about your healthcare experiences, education, and awareness of health initiatives.
+Your responses will remain completely anonymous, and no personal data will be shared or used beyond this research.
+This survey should take 5 to 7 minutes to complete.
+Thank you for taking the time to share your insights.
+
+**Survey Coordinator**: KarmenZhu\
+**Email**: karmen.zhu\@mail.utoronto.ca
+
+**Section 1: Personal Information**
+
+**1. Please select your age group:**(Select one)
+
+-   Under 20
+
+-   20-29
+
+-   30-39
+
+-   40-49
+
+-   50-59
+
+-   60 or older
+
+**2. What is your country of residence? **
+
+-  (please specify):\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_
+
+**3. What is your highest level of education completed or currently pursuing?** (Select one)
+
+-   No formal education
+
+-   Primary school
+
+-   High school diploma
+
+-   Some college/university
+
+-   Bachelor’s degree
+
+-   Master’s degree
+
+-   Doctoral degree (e.g., Ph.D.)
+
+-   Professional degree (e.g., MD, JD, MBA)
+
+-   Other (please specify): \_\_\_\_\_\_\_\_\_\_\_\_\_\_\_
+
+**4. What is your primary occupation?** (Select one)
+
+-   Full-time employed
+
+-   Part-time employed
+
+-   Self-employed
+
+-   Student
+
+-   Retired
+
+-   Other (please specify): \_\_\_\_\_\_\_\_\_\_\_\_\_\_\_
+
+**Section 2: Health and Healthcare Information**
+
+**5. Have you received a diphtheria immunization in the past 10 years?** (Select one)
+
+-   Yes
+
+-   No
+
+-   I don’t know
+
+**6. What is the approximate distance to the nearest healthcare facility (e.g., clinic, hospital)?** (Select one)
+
+-   Less than 5 km (3 miles)
+
+-   5-10 km (3-6 miles)
+
+-   11-20 km (7-12 miles)
+
+-   More than 20 km (12 miles)
+
+**7. How many walking-distance clinics or hospitals are located near your home?** (Select one)
+
+-   None
+
+-   1-2
+
+-   3-4
+
+-   5 or more
+
+**8. How long do you typically wait to receive care at a government healthcare facility?** (Select one)
+
+-   Less than 1 hour
+
+-   1-3 hours
+
+-   3-6 hours
+
+-   More than 6 hours
+
+**9. How much of your medical expenses (e.g., doctor visits, hospital stays, medication) does your health insurance cover?**
+
+-   (please specify):\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_
+
+**10. How much of your monthly income is spent on healthcare (including insurance, out-of-pocket expenses, etc.)?** (Select one)
+
+-   Less than 5%
+
+-   5-10%
+
+-   11-20%
+
+-   More than 20%
+
+**11. Do you have a family doctor? If so, how often do you visit your family doctor?** (Select one)
+
+-   never
+
+-   Rarely (Once a year)
+
+-   2–3 times a year
+
+-   More than 3 times a year
+
+**Final Section**
+
+Thank you for completing this survey! Your responses will help improve public health initiatives.
+
+## Additional Tables & Figures
+
+```{r}
+#| label: tbl-additionaltable
+#| tbl-cap: Summary of the Life Expectancy model, which includes key variables such as schooling years and the percentage of government health expenditure.
+The table presents the model coefficients along with their standard errors.
+#| echo: false
+#| eval: true
+#| warning: false
+#| message: false
+
+# Load the model
+lm_overall <- readRDS(here::here("models/lm_overall.rds"))
+
+# Extract summary statistics using broom
+coefficients_df <- tidy(lm_overall) %>%
+  mutate(
+    Term_Type = "Linear"  # Add a column specifying term type
+  )
+
+# Clean and format the Term column
+coefficients_df <- coefficients_df %>%
+  mutate(
+    term = gsub("[^[:alnum:]_\\s]", "", term)  # Remove special characters
+  )
+
+# Select only the required columns
+coefficients_cleaned <- coefficients_df %>%
+  select(term, estimate, std.error, statistic, p.value, Term_Type) %>%
+  rename(
+    Term = term,
+    Estimate = estimate,
+    `Std. Error` = std.error,
+    `t Value` = statistic,
+    `Pr(>|t|)` = p.value
+  )
+
+# Create the final table
+kable(
+  coefficients_cleaned,
+  format = "html",
+  col.names = c("Term", "Estimate", "Std. Error", "t Value", "Pr(>|t|)", "Term_Type"),
+  digits = 3,
+  caption = "Summary of the Life Expectancy model with key predictors and their contributions."
+)
+
+```
 

### 2024-12-04 other/shiny_app/LifeSpan_Predictor.R

@@ -1,0 +1,65 @@
+library(shiny)
+library(dplyr)
+library(shinyjs)
+
+# Fit the linear model
+lm_overall <- lm(LifeExpectancy ~ Status + Diphtheria + Schooling + TotalExpenditure * pct_doctor * pct_nursing, data = analysis_data)
+
+# Assuming data is already loaded and model is built
+data <- data %>%
+  mutate(Status = as.numeric(Status))
+
+# Define UI, set up default value
+ui <- fluidPage(
+  useShinyjs(),  # Initialize shinyjs for JavaScript functionality
+  titlePanel("Life Expectancy Prediction"),
+  
+  sidebarLayout(
+    sidebarPanel(
+      numericInput("status", "Status (Numeric):", value = 1, min = 0, max = 2),
+      numericInput("diphtheria", "Diphtheria Vaccination Rate (%):", value = 90, min = 0, max = 100),
+      numericInput("schooling", "Years of Schooling:", value = 13.14, min = 0, max = 20),
+      numericInput("total_expenditure", "Total Expenditure on Health (% of GDP):", value = 6.093, min = 0, max = 100),
+      numericInput("pct_doctor", "% of Population with Access to Doctors:", value = 19.45, min = 0, max = 100),
+      numericInput("pct_nursing", "% of Population with Access to Nurses:", value = 46.05, min = 0, max = 100),
+      
+      actionButton("predict", "Predict Life Expectancy")
+    ),
+    
+    mainPanel(
+      verbatimTextOutput("prediction_output"),
+      div(id = "loading", style = "display: none;", "Loading... Please wait")
+    )
+  )
+)
+
+# Define server logic
+server <- function(input, output) {
+  
+  observeEvent(input$predict, {  # Trigger prediction when the button is clicked
+    shinyjs::show("loading")  # Show loading message
+    
+    # Collect the input data into a data frame
+    input_data <- data.frame(
+      Status = as.character(input$status),
+      Diphtheria = as.numeric(input$diphtheria),
+      Schooling = as.numeric(input$schooling),
+      TotalExpenditure = as.numeric(input$total_expenditure),
+      pct_doctor = as.numeric(input$pct_doctor),
+      pct_nursing = as.numeric(input$pct_nursing)
+    )
+    
+    # Predict the life expectancy using the linear model
+    prediction <- predict(lm_overall, newdata = input_data)
+    
+    # Display the predicted life expectancy
[…]

### 2024-12-04 scripts/00-simulate_data.R

@@ -1,0 +1,71 @@
+#### Preamble ####
+# Purpose: Simulates a dataset of Life Expectancy
+# Author: Manjun Zhu
+# Date: 27 November 2024
+# Contact: karmen.zhu@mail.utoronto.ca
+# License: MIT
+# Pre-requisites: The `tidyverse` package must be installed
+# Any other information needed? Make sure you are in the `lifeexpectancy.Rproj`
+
+
+#### Workspace setup ####
+library(tidyverse)
+library(here)
+library(lubridate)
+library(arrow)
+
+set.seed(955)
+
+# Generate the number of rows for the dataset
+n <- 500
+
+# Simulated data
+# Generate some countries
+data <- function(n) {
+  country_names <- c("Australia", "Canada", "Germany", "Japan", "United States", 
+                     "Brazil", "India", "South Africa", "China", "Russia", 
+                     "France", "Italy", "United Kingdom", "South Korea", "Mexico", 
+                     "Argentina", "Spain", "Saudi Arabia", "Egypt", "Turkey", 
+                     "Nigeria", "Indonesia", "Malaysia", "Thailand", "Vietnam", 
+                     "Chile", "Poland", "Israel", "Pakistan", "Singapore", "Netherlands")
+  
+  # Randomly sample countries from the list
+  Country <- sample(country_names, n, replace = TRUE)
+  
+  # Randomly generate the data for each variable based on the summary statistics provided
+  Year <- sample(2004:2015, n, replace = TRUE)
+  
+  # Status (Developed/Developing) based on region or income classification
+  Status <- sample(c("Developed", "Developing"), n, replace = TRUE, prob = c(0.5, 0.5))
+  
+  # LifeExpectancy simulated around a mean of 72.24 with some variation
+  LifeExpectancy <- rnorm(n, mean = 72.24, sd = 10)
+  
+  # Diphtheria immunization rates between 4 and 99
+  Diphtheria <- round(runif(n, min = 4, max = 99), 2)
+  
+  # HealthExpenditure between 1.12 and 14.39
+  HealthExpenditure <- round(runif(n, min = 1.12, max = 14.39), 2)
+  
+  # Schooling years between 3.1 and 20.4
+  Schooling <- round(runif(n, min = 3.1, max = 20.4), 2)
+  
+  # doctor_per_10000 between 0.13 and 75.67
[…]

### 2024-12-04 scripts/01-test_simulated_data.R

@@ -1,0 +1,112 @@
+#### Preamble ####
+# Purpose: Tests the structure and validity of the simulated Life Expectancy
+# Author: Manjun Zhu
+# Date: 27 November 2024
+# Contact: karmen.zhu@mail.utoronto.ca
+# License: MIT
+# Pre-requisites: 
+  # - The `tidyverse` package must be installed and loaded
+  # - 00-simulate_data.R must have been run
+# Any other information needed? Make sure you are in the `lifeexpectancy.Rproj`
+
+
+#### Workspace setup ####
+library(tidyverse)
+library(here)
+
+data <- read_parquet(here::here("data/00-simulated_data/simulated_data.parquet"))
+
+
+# Test if the data was successfully loaded
+if (exists("data")) {
+  message("Test Passed: The dataset was successfully loaded.")
+} else {
+  stop("Test Failed: The dataset could not be loaded.")
+}
+
+#### Test data ####
+
+# Check if the dataset has the expected number of rows (e.g., 500 rows for the full dataset)
+if (nrow(data) == 500) {
+  message("Test Passed: The dataset has 500 rows.")
+} else {
+  stop("Test Failed: The dataset does not have 500 rows.")
+}
+
+# Check if the dataset has 9 columns
+if (ncol(data) == 9) {
+  message("Test Passed: The dataset has 9 columns.")
+} else {
+  stop("Test Failed: The dataset does not have 9 columns.")
+}
+
+# Check if all countries are valid and non-empty
+if (all(data$Country != "")) {
+  message("Test Passed: All countries have valid non-empty names.")
+} else {
+  stop("Test Failed: Some country names are missing or empty.")
+}
+
+# Check if the 'Status' column contains only "Developed" or "Developing"
+valid_statuses <- c("Developed", "Developing")
+if (all(data$Status %in% valid_statuses)) {
+  message("Test Passed: The 'Status' column contains only valid statuses.")
+} else {
+  stop("Test Failed: The 'Status' column contains invalid entries.")
+}
+
+# Check if the 'Year' column includes only values between 2004 and 2015
+if (all(data$Year >= 2004 & data$Year <= 2015)) {
+  message("Test Passed: The 'Year' column contains only valid years (2004-2015).")
+} else {
+  stop("Test Failed: The 'Year' column contains invalid years.")
+}
+
+# Check if there are no missing values in the dataset
+if (all(!is.na(data))) {
[…]

### 2024-12-04 scripts/02-download_data.R

@@ -1,0 +1,41 @@
+#### Preamble ####
+# Purpose: Downloads and saves the data from 'https://www.kaggle.com/datasets/kumarajarshi/life-expectancy-who'
+# Author: Manjun Zhu
+# Date: 16 November 2024
+# Contact: karmen.zhu@utoronto.ca
+# License: University of Toronto
+# Pre-requisites: No
+
+
+#### Workspace setup ####
+library(tidyverse)
+library(dplyr)
+
+#### Download data ####
+raw_data <- read.csv(file = "data/01-raw_data/LifeExpectancyData.csv")
+
+# Load the supplementary dataset
+doctor <- read.csv("data/01-raw_data/medicalDoctors.csv")
+nursing <- read.csv("data/01-raw_data/nursingAndMidwife.csv")
+
+# Ensure the column names align (modify as needed)
+# Example: rename 'Location' in doctor to 'Country'
+colnames(doctor)[colnames(doctor) == "Location"] <- "Country"
+colnames(doctor)[colnames(doctor) == "Period"] <- "Year"
+colnames(doctor)[colnames(doctor) == "First.Tooltip"] <- "pct_doctor"
+colnames(nursing)[colnames(nursing) == "Location"] <- "Country"
+colnames(nursing)[colnames(nursing) == "Period"] <- "Year"
+colnames(nursing)[colnames(nursing) == "First.Tooltip"] <- "pct_nursing"
+
+# Merge raw_data with doctor dataset
+test_data <- raw_data %>%
+  left_join(doctor %>% select(Country, Year, pct_doctor), by = c("Country", "Year"))
+
+# Merge test_data with nursing dataset
+raw_data <- test_data %>%
+  left_join(nursing %>% select(Country, Year, pct_nursing), by = c("Country", "Year"))
+
+
+#### Save data ####
+# change raw_data to whatever name you assigned when you downloaded it.
+write_csv(raw_data, "data/01-raw_data/raw_data.csv") 

### 2024-12-04 scripts/03-clean_data.R

@@ -1,0 +1,34 @@
+#### Preamble ####
+# Purpose: Clean the raw data recorded by WHO, removing unnecessary columns, and save it as csv and parquet.
+# Author: Manjun Zhu
+# Date: 23 November 2024
+# Contact: karmen.zhu@mail.utoronto.ca
+# License: MIT
+# Pre-requisites: finish raw-data
+
+#### Workspace setup ####
+library(tidyverse)
+library(arrow)
+
+#### Clean data ####
+data <- read_csv("data/01-raw_data/raw_data.csv")
+
+analysis_data <- data %>%
+  # Select the desired columns
+  select(Country, Year, Status, Life.expectancy, Diphtheria, 
+         Total.expenditure, Schooling, pct_doctor, pct_nursing) %>%
+  # Filter rows where Year is between 2009 and 2015, and drop the unrealistic value
+  filter(Year >= 2004 & Year <= 2015) %>%
+  filter(pct_doctor >= 0 & pct_doctor <= 1500) %>%
+  filter(pct_nursing >= 0 & pct_nursing <= 1500) %>%
+  # Remove rows with any missing values
+  drop_na()
+
+#### Rename column
+colnames(analysis_data)[colnames(analysis_data) == "Life.expectancy"] <- "LifeExpectancy"
+colnames(analysis_data)[colnames(analysis_data) == 
+                 "Total.expenditure"] <- "TotalExpenditure"
+
+#### Save data ####
+write_csv(analysis_data, "data/02-analysis_data/analysis_data.csv")
+write_parquet(analysis_data, "data/02-analysis_data/analysis_data.parquet")

### 2024-12-04 scripts/04-test_analysis_data.R

@@ -1,0 +1,112 @@
+#### Preamble ####
+# Purpose: Performs a series of validation tests on cleaned Life Expectancy Data
+# Author: Manjun Zhu
+# Date: 23 November 2024
+# Contact: karmen.zhu@mail.utoronto.ca
+# License: MIT
+# Pre-requisites: The `tidyverse` , `arrow`, and `testthat` packages must be installed.
+# Any other information needed? Make sure you are in the `lifeexpectancy.Rproj`
+
+
+#### Workspace setup ####
+library(tidyverse)
+library(testthat)
+library(arrow)
+
+analysis_data <- read_parquet(here::here("data/02-analysis_data/analysis_data.parquet"))
+
+
+#### Test data ####
+# Create list for results
+test_results <- list()
+
+# Test structure of data
+test_results$structure <- test_that("Data structure", {
+  expect_true(is.data.frame(analysis_data), 
+              info = "analysis_data is not a data frame.")
+  expect_equal(ncol(analysis_data), 10, info = "analysis_data does 
+               not have the correct number of columns.")
+  expect_equal(nrow(analysis_data), 1021, info = "analysis_data does 
+               not have the correct number of rows.")
+})
+
+# Test if column names match expected names
+test_results$column_names_test <- test_that("Column names", {
+  expected_colnames <- c("Country", "Year", "Status", "LifeExpectancy", "GDP", 
+                         "Diphtheria", "TotalExpenditure", "BMI", "Schooling", 
+                         "pct_doctor", "pct_nursing")
+  expect_equal(names(analysis_data), expected_colnames, 
+               info = "Column names do not match expected names.")
+})
+
+# Test if certain columns contain NA values
+test_results$na_values_test <- test_that("NA values", {
+  expect_false(any(is.na(analysis_data$LifeExpectancy)), 
+               info = "LifeExpectancy contains NA values.")
+  expect_false(any(is.na(analysis_data$Diphtheria)), 
+               info = "Diphtheria contains NA values.")
+  expect_false(any(is.na(analysis_data$pct_doctor)), 
+               info = "pct_doctor contains NA values.")
+  expect_false(any(is.na(analysis_data$pct_nursing)), 
+               info = "pct_nursing contains NA values.")
+  expect_false(any(is.na(analysis_data$Schooling)), 
+               info = "Schooling contains NA values.")
+  expect_false(any(is.na(analysis_data$TotalExpenditure)), 
[…]

### 2024-12-04 scripts/05-exploratory_data_analysis.R

@@ -1,0 +1,45 @@
+#### Preamble ####
+# Purpose: This script performs exploratory data analysis (EDA) on the cleaned life expectancy data
+# Author: Manjun Zhu
+# Date: 26 November 2024
+# Contact: karmen.zhu@mail.utoronto.ca
+# License: MIT
+# Pre-requisites: The 'tidyverse', 'here', and 'lubridate' packages must be installed
+
+
+# load libraries
+library(tidyverse)
+library(here)
+library(lubridate)
+
+# read cleaned data
+data <- read_parquet(here::here("data/02-analysis_data/analysis_data.parquet"))
+
+# Summary of numerical variables
+summary_stats <- data %>%
+  summarise(
+    LifeExpectancy = mean(LifeExpectancy, na.rm = TRUE),
+    Diphtheria = mean(Diphtheria, na.rm = TRUE),
+    TotalExpenditure = mean(TotalExpenditure, na.rm = TRUE),
+    BMI = mean(BMI, na.rm = TRUE),
+    Status = mean(as.numeric(Status), na.rm = TRUE),
+    Schooling = mean(Schooling, na.rm = TRUE),
+    pct_doctor = mean(pct_doctor, na.rm = TRUE),
+    pct_nursing = mean(pct_nursing, na.rm = TRUE),
+  )
+
+print(summary_stats)
+
+
+# Status breakdown
+status_distribution <- data %>%
+  count(Status) %>%
+  mutate(percent = n / sum(n) * 100)
+
+print(status_distribution)
+
+# number of NA values in each column
+missing_data <- data |>
+  summarise(across(everything(), ~ sum(is.na(.))))
+missing_data
+

### 2024-12-04 scripts/06-model_data.R

@@ -1,0 +1,50 @@
+#### Preamble ####
+# Purpose: This script performs out-of-sample testing and calculates the RMSE for the linear model trained in the previous script.
+# Author: Manjun Zhu
+# Date: 26 November 2024
+# Contact: karmen.zhu@mail.utoronto.ca
+# License: MIT
+# Pre-requisites: The `tidyverse`, `caret`, and 'here' packages must be installed
+
+
+#### Workspace setup ####
+library(tidyverse)
+library(here)
+
+#### Read data ####
+# convert some columns to the appropriate data types
+data <- read_parquet(here::here("data/02-analysis_data/analysis_data.parquet"))
+data <- data %>%
+  mutate(
+    Country = as.factor(Country),
+    Year = as.numeric(Year),
+    LifeExpectancy = as.numeric(LifeExpectancy),
+    Status = as.factor(Status),
+    Diphtheria = as.numeric(Diphtheria),
+    TotalExpenditure = as.numeric(TotalExpenditure),
+    Schooling = as.numeric(Schooling),
+    pct_doctor = as.numeric(pct_doctor),
+    pct_nursing = as.numeric(pct_nursing)
+  )
+
+### Model data ####
+# linear model for Developed Countries' Life Expectancy based on socioeconomic factors
+lm_developed <- lm(LifeExpectancy ~  Diphtheria + Schooling + TotalExpenditure * pct_doctor * pct_nursing, 
+                   data = data %>% filter(Status == "Developed"))
+
+# linear model for Developing Countries' Life Expectancy based on socioeconomic factors
+lm_developing <- lm(LifeExpectancy ~ Diphtheria + Schooling + TotalExpenditure * pct_doctor * pct_nursing, 
+                    data = data %>% filter(Status == "Developing"))
+
+# linear model for overall Life Expectancy based on socioeconomic factors
+# Since Status is concluded in this model, first need to make it be numeric
+data <- data %>%
+  mutate(Status = as.numeric(Status))
+lm_overall <- lm(LifeExpectancy ~ Status + Diphtheria + Schooling + TotalExpenditure * pct_doctor * pct_nursing, 
+                 data = data)
+
+#### Save model ####
+saveRDS(lm_developed, here::here("models/lm_developed.rds"))
+saveRDS(lm_developing, here::here("models/lm_developing.rds"))
+saveRDS(lm_developing, here::here("models/lm_overall.rds"))
+

### 2024-12-04 scripts/07-replications.R

@@ -1,0 +1,53 @@
+#### Preamble ####
+# Purpose: This script performs out-of-sample testing for the model that was presented in the 06-model_data
+# Author: Manjun Zhu
+# Date: 23 November 2024
+# Contact: karmen.zhu@mail.utoronto.ca
+# License: MIT
+# Pre-requisites: Ensure the `tidyverse`, `caret`, and `here` packages are installed.
+# Any other information needed? Make sure you are in the `lifeexpectancy.Rproj`
+
+
+#### Workspace setup ####
+library(tidyverse)
+library(caret)
+library(here)
+
+# Load the cleaned dataset (adjust the file path as needed)
+data <- read_parquet(here::here("data/02-analysis_data/analysis_data.parquet"))
+
+# Clean the data again to ensure consistency with the model
+data_cleaned <- data %>%
+  filter(!is.na(LifeExpectancy), !is.na(Diphtheria), !is.na(Schooling), 
+         !is.na(TotalExpenditure), !is.na(pct_doctor)) %>%
+  mutate(
+    LifeExpectancy = as.numeric(LifeExpectancy),
+    Diphtheria = as.numeric(Diphtheria),
+    Schooling = as.integer(Schooling),
+    TotalExpenditure = as.integer(TotalExpenditure),
+    pct_doctor = as.integer(pct_doctor)
+  )
+
+# Split the data into training (80%) and testing (20%) sets
+set.seed(955) # For reproducibility
+Index <- createDataPartition(data_cleaned$LifeExpectancy,
+                                  p = 0.8,
+                                  list = FALSE,
+                                  times = 1
+)
+
+train_data <- data_cleaned[Index, ]
+test_data <- data_cleaned[-Index, ]
+
+# Load the saved model
+finisher_model <- readRDS(here::here("models/lm_developed.rds"))
+
+# Perform out-of-sample prediction on the test set
+predicted <- predict(finisher_model, newdata = test_data)
+
+# Calculate the RMSE (Root Mean Squared Error)
+rmse <- sqrt(mean((predicted - test_data$LifeExpectancy)^2))
+
+# Print RMSE value
+cat("Root Mean Squared Error (RMSE) on the test data: ", round(rmse, 2), "\n")
+

### 2024-12-04 paper/paper.qmd
Add files via upload

@@
 format: pdf
-toc: true
-toc-depth: 3
 number-sections: true

### 2024-12-04 paper/paper.qmd
Add files via upload

--- 2024-12-04 paper/paper.qmd
@@
 
-For the developed countries model, predictors like Diphtheria Immunization, Schooling, Total Expenditure, and the number of Doctors per 10,000 were included.
+For the developed countries model, predictors like Diphtheria Immunization, Schooling, Health Expenditure, and the number of Doctors per 10,000 were included.
 These variables are well-established factors influencing life expectancy in high-income nations [@life_expectancy_who; @world_health_statistics].
@@
 #| label: tbl-keyestimates
-#| tbl-cap: "Summary of key model estimates for Developed and Developing counties, including coefficients for predictors like Diphtheria, Schooling, and TotalExpenditure, pct_doctor, pct_nursing, with standard errors for each estimate.
+#| tbl-cap: "Summary of key model estimates for Developed and Developing counties, including coefficients for predictors like Diphtheria, Schooling, and Percentage of HealthExpenditure, pct_doctor, pct_nursing, with standard errors for each estimate.
 Model performance statistics, such as sample size, R², and adjusted R², are also displayed."
@@
 A comparison of predicted life expectancy distributions reveals several notable trends.
-Longer schooling years[@fig-schoolgroups] strongly correlate with higher life expectancy, highlighting the critical role of education.
-High diphtheria immunization coverage[@fig-diphtheria-groups] generally predicts greater life expectancy, though medium and low coverage levels exhibit variability, suggesting that immunization is not the sole or dominant factor.
-Similarly, higher government health expenditure[@fig-totalexpenditure-groups] aligns with improved outcomes, while lower spending shows minimal variation.
-Availability of doctors and nurses[@fig-doctornursing-groups] also indicates a positive relationship with life expectancy, though some variability suggests other contributing factors are at play.
+Longer schooling years strongly correlate with higher life expectancy, highlighting the critical role of education.
+High diphtheria immunization coverage generally predicts greater life expectancy, though medium and low coverage levels exhibit variability, suggesting that immunization is not the sole or dominant factor.
+Similarly, higher percentage of government health expenditure aligns with improved outcomes, while lower spending shows minimal variation.
+Availability of doctors and nurses also indicates a positive relationship with life expectancy, though some variability suggests other contributing factors are at play.
 
